---
title: "Deconvolution Diagnostics & Signature Matrix Validation"
subtitle: "Pipeline step 4 of 4 -- run after 1_TG_sc_reprocessing.qmd and 3_TG_incoming_sample_QC_maturation.qmd; interrogates why NNLS collapses to ~2 identities"
author: "Austin Hovland"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Purpose

The weekly maturation notebook (`3_TG_incoming_sample_QC_maturation.qmd`) flags that its NNLS deconvolution output is suspicious: nearly all TG-bulk specimen-weeks collapse onto essentially **two** reference identities (a dominant NF2 fraction plus one non-neuronal identity), with every other column in the 11-identity signature matrix reading exactly 0.0. That is either (a) a genuine and striking biological result, or (b) a signature matrix that is not mathematically identifiable at the resolution the plots imply. This notebook does not re-run the weekly QC; it exists purely to adjudicate that question with four diagnostics run in order:

1.  **Signature matrix collinearity audit** -- pairwise correlation and variance-inflation-factor (VIF) analysis of the 11 reference identity columns, restricted to the exact marker genes used by NNLS.
2.  **Synthetic mixture recovery test** -- known synthetic mixtures built from the reference profiles themselves are passed back through the identical NNLS pipeline; failure to recover ground truth here proves an identifiability problem independent of any biology in the real samples.
3.  **Alternative deconvolution cross-validation** -- the same incoming samples are deconvolved with a second, differently-regularized method (ridge-dampened least squares) and compared against the NNLS fractions.
4.  **Marker-gene specificity re-screen** -- every curated marker gene is ranked by fold-change specificity against its next-best-matching identity, surfacing markers that are actually shared across subtypes and therefore contributing to collapse.

This notebook assumes `1_TG_sc_reprocessing.qmd` has been run (so `results/pseudobulk/` is populated) and that `3_TG_incoming_sample_QC_maturation.qmd` has been run at least once (so `results/incoming_qc/` contains the current incoming-sample joint-normalized data and NNLS output referenced below).

## Setup

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(edgeR)

if (!requireNamespace("nnls", quietly = TRUE)) {
  install.packages("nnls", repos = "https://cloud.r-project.org")
}
library(nnls)

dir.create("results/deconvolution_diagnostics", recursive = TRUE, showWarnings = FALSE)

combined_dge <- readRDS(
  "results/pseudobulk/GSE197289_TG_combined_reference_edgeR_dge.rds"
)

combined_logcpm <- as.matrix(
  read.csv(
    "results/pseudobulk/GSE197289_TG_combined_reference_logCPM.csv",
    row.names = 1,
    check.names = FALSE
  )
)

combined_metadata <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_reference_metadata.csv",
  show_col_types = FALSE
)

combined_marker_manifest <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_marker_manifest.csv",
  show_col_types = FALSE
)

incoming_sample_metadata <- read_csv(
  "results/incoming_qc/incoming_sample_metadata.csv",
  show_col_types = FALSE
)

stopifnot(identical(colnames(combined_logcpm), combined_metadata$pseudobulk_id))

neuronal_identity_labels <- combined_metadata |>
  filter(reference_level == "Neuronal subtype") |>
  pull(reference_label) |>
  as.character() |>
  unique()
```

## Rebuild the exact signature matrix used by NNLS

To diagnose the deconvolution honestly, this section reconstructs the identical identity-averaged signature matrix that `3_TG_incoming_sample_QC_maturation.qmd` passes into `nnls::nnls()` -- same marker genes, same per-identity averaging -- rather than a fresh approximation, so every downstream diagnostic is evaluating the actual object responsible for the striking result.

```{r}
#| label: rebuild-signature-matrix

deconv_signature_genes <- combined_marker_manifest |>
  filter(gene %in% rownames(combined_logcpm)) |>
  pull(gene) |>
  unique()

signature_matrix_linear <- 2^combined_logcpm[deconv_signature_genes, , drop = FALSE]

signature_identity_map <- combined_metadata |>
  transmute(pseudobulk_id, reference_label = as.character(reference_label)) |>
  filter(pseudobulk_id %in% colnames(signature_matrix_linear))

identity_levels <- unique(signature_identity_map$reference_label)

signature_matrix_by_identity <- sapply(identity_levels, function(id_lab) {
  cols <- signature_identity_map$pseudobulk_id[signature_identity_map$reference_label == id_lab]
  if (length(cols) == 1) return(signature_matrix_linear[, cols])
  rowMeans(signature_matrix_linear[, cols, drop = FALSE])
})
colnames(signature_matrix_by_identity) <- identity_levels

signature_summary <- tibble(
  metric = c(
    "Marker genes in signature matrix",
    "Reference identities (columns)",
    "Neuronal-subtype identities",
    "Broad non-neuronal identities"
  ),
  value = c(
    nrow(signature_matrix_by_identity),
    ncol(signature_matrix_by_identity),
    sum(identity_levels %in% neuronal_identity_labels),
    sum(!identity_levels %in% neuronal_identity_labels)
  )
)

knitr::kable(signature_summary)
```

## Diagnostic 1 -- signature matrix collinearity audit

If NF1, NF2, NF3, PEP, and NP are highly correlated with each other across the curated marker genes, NNLS has no mathematical way to assign expression uniquely to one of them -- it will arbitrarily favor whichever column best explains the residual, typically the one with marginally higher marker expression on average. The condition number and per-identity VIF quantify exactly how ill-posed this system is, independent of any real sample.

```{r}
#| label: fig-signature-collinearity
#| fig-width: 10
#| fig-height: 9

log_signature_matrix <- log2(signature_matrix_by_identity + 1)

identity_cor <- cor(log_signature_matrix, method = "pearson")

pheatmap(
  identity_cor,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  number_format = "%.2f",
  main = "Pairwise correlation between reference identity signature columns\n(log2 marker-gene expression, the exact NNLS signature matrix)"
)

compute_vif <- function(x) {
  vif_out <- map_dbl(seq_len(ncol(x)), function(i) {
    y <- x[, i]
    preds <- x[, -i, drop = FALSE]
    fit <- lm(y ~ preds)
    r2 <- summary(fit)$r.squared
    if (r2 >= 0.999999) return(Inf)
    1 / (1 - r2)
  })
  names(vif_out) <- colnames(x)
  vif_out
}

identity_vif <- compute_vif(log_signature_matrix)

vif_table <- tibble(
  reference_identity = names(identity_vif),
  vif = identity_vif
) |>
  mutate(
    collinearity_flag = case_when(
      vif >= 10 ~ "severe (VIF >= 10)",
      vif >= 5 ~ "moderate (VIF 5-10)",
      TRUE ~ "acceptable (VIF < 5)"
    )
  ) |>
  arrange(desc(vif))

knitr::kable(vif_table, digits = 2)

signature_matrix_condition_number <- kappa(log_signature_matrix, exact = TRUE)

message(
  "Signature matrix condition number: ", round(signature_matrix_condition_number, 1),
  " (values above ~30 are generally considered ill-conditioned for regression)"
)

write.csv(
  vif_table,
  "results/deconvolution_diagnostics/signature_matrix_vif.csv",
  row.names = FALSE
)

write.csv(
  as.data.frame(identity_cor),
  "results/deconvolution_diagnostics/signature_matrix_identity_correlation.csv"
)
```

A VIF above 10 (or a very high condition number) for NF1/NF2/NF3/PEP/NP would directly explain why the solver collapses to one dominant neuronal column regardless of the true underlying mixture -- the columns are not statistically distinguishable given this marker gene set.

## Diagnostic 2 -- synthetic mixture recovery test

This is the most direct test available: build known mixtures **from the reference pseudobulk profiles themselves**, run them through the identical NNLS call used on real incoming samples, and check whether the true mixing proportions come back out. If a 100% single-identity spike-in already fails to recover cleanly, or a multi-identity mixture collapses to fewer identities than were mixed in, the problem is the signature matrix itself, not the biology of any real sample.

```{r}
#| label: synthetic-recovery-test

run_nnls_fraction <- function(sample_expr, sig_matrix) {
  fit <- nnls::nnls(as.matrix(sig_matrix), as.numeric(sample_expr))
  coefs <- pmax(fit$x, 0)
  if (sum(coefs) == 0) coefs <- rep(0, length(coefs))
  else coefs <- coefs / sum(coefs)
  names(coefs) <- colnames(sig_matrix)
  coefs
}

set.seed(197289)

build_synthetic_mixture <- function(true_props, sig_matrix) {
  stopifnot(all(names(true_props) %in% colnames(sig_matrix)))
  full_props <- setNames(rep(0, ncol(sig_matrix)), colnames(sig_matrix))
  full_props[names(true_props)] <- true_props
  as.numeric(sig_matrix %*% full_props)
}

synthetic_scenarios <- list(
  pure_NF2 = c(NF2 = 1),
  pure_NF1 = c(NF1 = 1),
  pure_PEP = c(PEP = 1),
  two_way_NF2_glia = c(NF2 = 0.6, `Satellite glia` = 0.4),
  three_way_realistic = c(NF2 = 0.5, NF3 = 0.2, `Satellite glia` = 0.3),
  five_neuronal_even = c(NF1 = 0.2, NF2 = 0.2, NF3 = 0.2, PEP = 0.2, NP = 0.2),
  neuronal_glia_immune = c(NF2 = 0.5, `Satellite glia` = 0.3, Immune = 0.2)
)

synthetic_scenarios <- synthetic_scenarios[
  map_lgl(synthetic_scenarios, function(p) all(names(p) %in% colnames(signature_matrix_by_identity)))
]

synthetic_results <- imap_dfr(synthetic_scenarios, function(true_props, scenario_name) {
  synthetic_expr <- build_synthetic_mixture(true_props, signature_matrix_by_identity)
  recovered <- run_nnls_fraction(synthetic_expr, signature_matrix_by_identity)

  true_full <- setNames(rep(0, length(recovered)), names(recovered))
  true_full[names(true_props)] <- true_props

  tibble(
    scenario = scenario_name,
    reference_identity = names(recovered),
    true_fraction = true_full[names(recovered)],
    recovered_fraction = recovered
  )
})

synthetic_summary <- synthetic_results |>
  group_by(scenario) |>
  summarise(
    n_true_identities = sum(true_fraction > 0),
    n_recovered_identities = sum(recovered_fraction > 0.01),
    max_abs_error = max(abs(true_fraction - recovered_fraction)),
    correctly_dominant = names(which.max(setNames(recovered_fraction, reference_identity)))[1],
    .groups = "drop"
  )

knitr::kable(synthetic_summary, digits = 3)

ggplot(
  synthetic_results |> filter(true_fraction > 0 | recovered_fraction > 0.01),
  aes(x = reference_identity, y = recovered_fraction, fill = reference_identity)
) +
  geom_col() +
  geom_point(aes(y = true_fraction), shape = 4, size = 3, stroke = 1.4, color = "black") +
  facet_wrap(~ scenario, scales = "free_x", ncol = 3) +
  labs(
    title = "Synthetic mixture recovery: bars = NNLS-recovered fraction, X = ground truth",
    subtitle = "Mixtures built directly from reference pseudobulk profiles and passed through the identical NNLS call",
    x = "Reference identity",
    y = "Fraction"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

write.csv(
  synthetic_results,
  "results/deconvolution_diagnostics/synthetic_mixture_recovery.csv",
  row.names = FALSE
)
```

If `n_recovered_identities` is consistently lower than `n_true_identities` across scenarios -- especially for `five_neuronal_even`, which mixes five neuronal subtypes in genuinely equal proportion -- that is direct proof the signature matrix cannot resolve those identities, independent of anything biological about TGN1003A, TGN1028B, TGN1215AvX, or TGN251028B.

## Diagnostic 3 -- alternative deconvolution cross-validation

NNLS has no regularization and will happily drive collinear coefficients to exactly zero rather than splitting credit between them, which can visually exaggerate a "two-identity" result. A ridge-dampened (L2-penalized) least-squares solve, re-normalized to sum to 1, handles collinearity differently -- comparing the two methods on the same real incoming samples shows whether the NF2-dominant pattern is a property of the biology or an artifact of the specific solver's handling of collinear columns.

```{r}
#| label: ridge-deconvolution-comparison
#| fig-width: 11
#| fig-height: 6

incoming_qc_flags <- read_csv(
  "results/incoming_qc/incoming_sample_qc_flags.csv",
  show_col_types = FALSE
)

incoming_deconv_existing <- read_csv(
  "results/incoming_qc/weekly_nnls_deconvolution_fractions.csv",
  show_col_types = FALSE
)

incoming_counts_path <- "RealTGN.salmon.merged.gene_counts.tsv"

incoming_quant <- read_tsv(incoming_counts_path, show_col_types = FALSE)
incoming_sample_ids_raw <- setdiff(colnames(incoming_quant), c("gene_id", "gene_name"))

incoming_counts_tbl <- incoming_quant |>
  transmute(
    gene_name = toupper(trimws(gene_name)),
    across(all_of(incoming_sample_ids_raw), as.numeric)
  ) |>
  filter(!is.na(gene_name), gene_name != "") |>
  group_by(gene_name) |>
  summarise(across(all_of(incoming_sample_ids_raw), \(x) sum(x, na.rm = TRUE)), .groups = "drop")

incoming_counts <- incoming_counts_tbl |> as.data.frame()
rownames(incoming_counts) <- incoming_counts$gene_name
incoming_counts$gene_name <- NULL
incoming_counts <- as.matrix(incoming_counts)

reference_gene_key <- toupper(rownames(combined_dge$counts))
shared_gene_key <- intersect(reference_gene_key, rownames(incoming_counts))

reference_counts_shared <- combined_dge$counts[match(shared_gene_key, reference_gene_key), , drop = FALSE]
incoming_counts_shared <- incoming_counts[match(shared_gene_key, rownames(incoming_counts)), , drop = FALSE]
rownames(reference_counts_shared) <- shared_gene_key
rownames(incoming_counts_shared) <- shared_gene_key

joint_counts <- cbind(reference_counts_shared, incoming_counts_shared)
joint_dge <- DGEList(counts = joint_counts)
joint_dge <- calcNormFactors(joint_dge, method = "TMM")
joint_logcpm <- cpm(joint_dge, log = TRUE, prior.count = 2)

incoming_ids <- incoming_sample_metadata$sample_id
incoming_linear_expr <- 2^joint_logcpm[deconv_signature_genes, incoming_ids, drop = FALSE]

run_ridge_fraction <- function(sample_expr, sig_matrix, lambda = 1) {
  A <- as.matrix(sig_matrix)
  b <- as.numeric(sample_expr)
  ridge_solution <- solve(t(A) %*% A + lambda * diag(ncol(A)), t(A) %*% b)
  coefs <- pmax(as.numeric(ridge_solution), 0)
  if (sum(coefs) == 0) coefs <- rep(0, length(coefs)) else coefs <- coefs / sum(coefs)
  names(coefs) <- colnames(sig_matrix)
  coefs
}

ridge_fractions <- apply(incoming_linear_expr, 2, run_ridge_fraction, sig_matrix = signature_matrix_by_identity, lambda = 1)
ridge_fractions <- as.data.frame(t(ridge_fractions)) |> rownames_to_column("sample_id")

ridge_long <- ridge_fractions |>
  pivot_longer(-sample_id, names_to = "reference_identity", values_to = "ridge_fraction") |>
  left_join(
    incoming_sample_metadata |> select(sample_id, specimen_id, sample_class, maturation_week),
    by = "sample_id"
  ) |>
  filter(!is.na(maturation_week)) |>
  group_by(specimen_id, sample_class, maturation_week, reference_identity) |>
  summarise(ridge_fraction = mean(ridge_fraction, na.rm = TRUE), .groups = "drop")

method_comparison <- incoming_deconv_existing |>
  rename(nnls_fraction = estimated_fraction) |>
  full_join(
    ridge_long,
    by = c("specimen_id", "sample_class", "maturation_week", "reference_identity")
  ) |>
  mutate(
    nnls_fraction = replace_na(nnls_fraction, 0),
    ridge_fraction = replace_na(ridge_fraction, 0),
    abs_disagreement = abs(nnls_fraction - ridge_fraction)
  )

method_agreement_summary <- method_comparison |>
  group_by(specimen_id, maturation_week) |>
  summarise(
    n_identities_nnls = sum(nnls_fraction > 0.01),
    n_identities_ridge = sum(ridge_fraction > 0.01),
    mean_abs_disagreement = mean(abs_disagreement),
    .groups = "drop"
  )

knitr::kable(method_agreement_summary, digits = 3)

ggplot(
  method_comparison |> filter(nnls_fraction > 0.01 | ridge_fraction > 0.01),
  aes(x = nnls_fraction, y = ridge_fraction, color = reference_identity)
) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 2.6, alpha = 0.85) +
  labs(
    title = "NNLS vs. ridge-dampened deconvolution fraction per identity",
    subtitle = "Points on the dashed line indicate method agreement; systematic deviation flags solver-dependent collapse",
    x = "NNLS-estimated fraction",
    y = "Ridge-estimated fraction (lambda = 1)",
    color = "Reference identity"
  ) +
  theme_classic()

write.csv(
  method_comparison,
  "results/deconvolution_diagnostics/nnls_vs_ridge_comparison.csv",
  row.names = FALSE
)
```

If the ridge solution consistently spreads mass across NF1/NF2/NF3/PEP/NP where NNLS puts everything on NF2 alone, that is strong evidence the true signal is "broadly neuronal, subtype-ambiguous" rather than "specifically NF2" -- an important distinction for how you describe maturation identity in any downstream report.

## Diagnostic 4 -- marker-gene specificity re-screen

The curated marker manifest was built by domain knowledge, not by re-checking specificity against this exact combined reference. This re-screen ranks every marker gene by how well it distinguishes its assigned identity from the single next-best-matching identity in the reference, surfacing markers that are quietly shared across subtypes and are therefore contributing degenerate columns to the signature matrix above.

```{r}
#| label: fig-marker-specificity-rescreen
#| fig-width: 12
#| fig-height: 8

marker_specificity <- combined_marker_manifest |>
  filter(gene %in% rownames(combined_logcpm)) |>
  rowwise() |>
  mutate(
    own_identity_expr = {
      own_cols <- signature_identity_map$pseudobulk_id[signature_identity_map$reference_label == marker_group]
      if (length(own_cols) == 0) NA_real_ else mean(combined_logcpm[gene, own_cols])
    }
  ) |>
  ungroup()

expr_by_identity_long <- combined_logcpm[unique(marker_specificity$gene), , drop = FALSE] |>
  as.data.frame() |>
  rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "pseudobulk_id", values_to = "log_cpm") |>
  left_join(signature_identity_map, by = "pseudobulk_id") |>
  group_by(gene, reference_label) |>
  summarise(mean_log_cpm = mean(log_cpm), .groups = "drop")

specificity_scores <- marker_specificity |>
  select(gene, marker_group) |>
  distinct() |>
  left_join(expr_by_identity_long, by = "gene") |>
  group_by(gene, marker_group) |>
  summarise(
    own_expr = mean_log_cpm[reference_label == marker_group][1],
    next_best_identity = reference_label[reference_label != marker_group][
      which.max(mean_log_cpm[reference_label != marker_group])
    ],
    next_best_expr = max(mean_log_cpm[reference_label != marker_group]),
    .groups = "drop"
  ) |>
  mutate(
    specificity_log2fc = own_expr - next_best_expr,
    low_specificity_flag = specificity_log2fc < 1
  ) |>
  arrange(specificity_log2fc)

knitr::kable(
  specificity_scores |> filter(low_specificity_flag) |> arrange(specificity_log2fc),
  digits = 2,
  caption = "Curated markers with < 2-fold specificity over their next-best-matching identity"
)

ggplot(
  specificity_scores,
  aes(x = specificity_log2fc, y = reorder(gene, specificity_log2fc), fill = marker_group)
) +
  geom_col() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  labs(
    title = "Marker-gene specificity vs. next-best-matching identity",
    subtitle = "Dashed line = 2-fold specificity floor; genes left of the line are the likely source of signature-matrix collinearity",
    x = "log2(own identity expression / next-best identity expression)",
    y = NULL,
    fill = "Marker program"
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(size = 7))

write.csv(
  specificity_scores,
  "results/deconvolution_diagnostics/marker_specificity_rescreen.csv",
  row.names = FALSE
)
```

Any marker gene flagged here as low-specificity is a direct, actionable target: either drop it from the signature matrix or replace it with a more specific alternative from the single-cell reference, then re-run `3_TG_incoming_sample_QC_maturation.qmd`'s NNLS section to see whether subtype resolution improves.

## Summary and verdict

```{r}
#| label: diagnostics-summary

diagnostics_summary <- tibble(
  diagnostic = c(
    "Max VIF among neuronal-subtype identities",
    "Signature matrix condition number",
    "Synthetic scenarios with recovered identity count < true identity count",
    "Mean |NNLS - ridge| disagreement across specimen-weeks",
    "Curated markers below 2-fold specificity floor"
  ),
  value = c(
    round(max(vif_table$vif[vif_table$reference_identity %in% neuronal_identity_labels], na.rm = TRUE), 1),
    round(signature_matrix_condition_number, 1),
    sum(synthetic_summary$n_recovered_identities < synthetic_summary$n_true_identities),
    round(mean(method_agreement_summary$mean_abs_disagreement), 3),
    sum(specificity_scores$low_specificity_flag)
  )
)

knitr::kable(diagnostics_summary)
```

Read this table before trusting any subtype-level NNLS fraction reported in `3_TG_incoming_sample_QC_maturation.qmd`. A high VIF/condition number, failed synthetic recovery, large NNLS-vs-ridge disagreement, and multiple low-specificity markers together would confirm that the "two-identity" result is a signature-matrix identifiability artifact; if instead these diagnostics come back clean, the NF2-dominant pattern deserves to be reported as a genuine biological finding rather than caveated as exploratory.

## Next steps checklist

- If VIF/condition-number flags are severe, collapse the signature matrix to fewer, well-separated super-identities (e.g. merge NF1/NF2/NF3 into one "neurofilament-high" column) before attempting subtype-resolution claims.
- If synthetic recovery fails even for `five_neuronal_even`, do not report subtype-level composition at all until the marker panel is rebuilt with identity-specific genes surfaced by Diagnostic 4.
- If NNLS and ridge disagree substantially, prefer the ridge (or another regularized) estimate for any external-facing report, since it degrades more gracefully under collinearity.
- Any marker gene flagged as low-specificity here should be cross-checked against the original GSE197289 subtype marker literature before removal, in case the shared expression reflects genuine biological overlap rather than a curation error.
- Re-run this notebook whenever the marker manifest or combined pseudobulk reference is rebuilt in `1_TG_sc_reprocessing.qmd`, since all four diagnostics depend directly on those artifacts.
