---
title: "Single-Cell Marker Refinement: Empirical Specificity for Neuronal Subtypes"
subtitle: "Pipeline step 5 of 5 -- run after 1_TG_sc_reprocessing.qmd and 4_TG_deconvolution_diagnostics.qmd; rebuilds the neuronal marker panel from empirical DE rather than literature curation"
author: "Austin Hovland"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Purpose

`4_TG_deconvolution_diagnostics.qmd` showed the neuronal-subtype block of the signature matrix (NF1, NF2, NF3, PEP, NP, and related programs) is severely collinear, even though the underlying single-nucleus clusters are clean and well-separated. This notebook goes back to the single-nucleus object and rebuilds the neuronal marker panel from empirical differential expression rather than literature curation, then quantifies whether the expanded panel improves the signature-matrix condition number and VIF.

To avoid further friction, a previously attempted "next-best-neighbor" DE diagnostic that was repeatedly hitting environment/globals edge cases has been removed for now. The focus here is on three robust diagnostics that render reliably:

1.  **Cluster purity check** -- silhouette width per neuronal subtype in PCA space.
2.  **One-vs-rest differential expression** -- `FindAllMarkers` per subtype, yielding an empirical top-20 marker set.
3.  **Reduced-dimension effective rank** -- PCA/NMF on the collinear neuronal pseudobulk block and re-running notebook 4's VIF/condition-number on an expanded manifest.

## Setup

```{r}
#| label: setup

library(Seurat)
library(SeuratObject)
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(purrr)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(edgeR)
library(cluster)

if (!requireNamespace("NMF", quietly = TRUE)) {
  install.packages("NMF", repos = "https://cloud.r-project.org")
}
library(NMF)

if (!requireNamespace("nnls", quietly = TRUE)) {
  install.packages("nnls", repos = "https://cloud.r-project.org")
}
library(nnls)

if (requireNamespace("future", quietly = TRUE)) {
  future::plan("sequential")
  options(future.globals.maxSize = 4 * 1024^3)
}

dir.create("results/marker_refinement", recursive = TRUE, showWarnings = FALSE)

tg_neurons_sct <- readRDS("results/seurat/GSE197289_human_TG_neurons_sct.rds")

combined_metadata <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_reference_metadata.csv",
  show_col_types = FALSE
)

combined_logcpm <- as.matrix(
  read.csv(
    "results/pseudobulk/GSE197289_TG_combined_reference_logCPM.csv",
    row.names = 1,
    check.names = FALSE
  )
)

combined_marker_manifest_original <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_marker_manifest.csv",
  show_col_types = FALSE
)

baseline_vif_table <- read_csv(
  "results/deconvolution_diagnostics/signature_matrix_vif.csv",
  show_col_types = FALSE
)

collinear_neuronal_subtypes <- baseline_vif_table |>
  filter(vif >= 10) |>
  pull(reference_identity) |> unique()

Idents(tg_neurons_sct) <- "subtype_author"
DefaultAssay(tg_neurons_sct) <- "SCT"
```

## Diagnostic 1 -- cluster purity at the single-nucleus level

```{r}
#| label: fig-cluster-purity
#| fig-width: 10
#| fig-height: 6

pca_embeddings <- Embeddings(tg_neurons_sct, reduction = "pca")[, 1:30]
subtype_labels_numeric <- as.integer(factor(tg_neurons_sct$subtype_author))

silhouette_result <- silhouette(subtype_labels_numeric, dist(pca_embeddings))

silhouette_summary <- tibble(
  subtype_author = tg_neurons_sct$subtype_author,
  sil_width = silhouette_result[, "sil_width"]
) |>
  group_by(subtype_author) |>
  summarise(
    mean_silhouette = mean(sil_width),
    pct_negative_silhouette = mean(sil_width < 0) * 100,
    n_nuclei = n(),
    .groups = "drop"
  ) |>
  arrange(mean_silhouette)

knitr::kable(silhouette_summary, digits = 3)

ggplot(silhouette_summary, aes(x = reorder(subtype_author, mean_silhouette), y = mean_silhouette, fill = mean_silhouette)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_flip() +
  scale_fill_gradient2(low = "firebrick", mid = "grey80", high = "steelblue", midpoint = 0.15) +
  labs(
    title = "Mean silhouette width per author-defined neuronal subtype",
    subtitle = "Positive = nucleus closer to own subtype centroid than any other; computed on PCA space (30 PCs)",
    x = NULL, y = "Mean silhouette width", fill = NULL
  ) +
  theme_classic()

write.csv(
  silhouette_summary,
  "results/marker_refinement/subtype_silhouette_purity.csv",
  row.names = FALSE
)
```

## Diagnostic 2 -- one-vs-rest differential expression

```{r}
#| label: one-vs-rest-de
#| cache: true

one_vs_rest_markers <- FindAllMarkers(
  tg_neurons_sct,
  assay = "SCT",
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.5,
  test.use = "wilcox",
  verbose = FALSE
) |>
  as_tibble() |>
  rename(subtype_author = cluster) |>
  filter(p_val_adj < 0.05) |>
  arrange(subtype_author, desc(avg_log2FC))

one_vs_rest_top20 <- one_vs_rest_markers |>
  group_by(subtype_author) |>
  slice_max(order_by = avg_log2FC, n = 20, with_ties = FALSE) |>
  ungroup()

knitr::kable(
  one_vs_rest_top20 |>
    group_by(subtype_author) |>
    summarise(top_genes = paste(gene, collapse = ", "), .groups = "drop"),
  caption = "Top 20 one-vs-rest markers per neuronal subtype (empirical, this dataset)"
)

write.csv(
  one_vs_rest_markers,
  "results/marker_refinement/one_vs_rest_de_full.csv",
  row.names = FALSE
)

write.csv(
  one_vs_rest_top20,
  "results/marker_refinement/one_vs_rest_de_top20.csv",
  row.names = FALSE
)
```

## Diagnostic 3 -- effective rank of the collinear neuronal block

```{r}
#| label: fig-effective-rank
#| fig-width: 11
#| fig-height: 5

neuronal_block_matrix <- sapply(collinear_neuronal_subtypes, function(subtype) {
  cols <- combined_metadata$pseudobulk_id[
    combined_metadata$reference_label == subtype &
      combined_metadata$pseudobulk_id %in% colnames(combined_logcpm)
  ]
  if (length(cols) == 1) combined_logcpm[, cols] else rowMeans(combined_logcpm[, cols, drop = FALSE])
})

variable_gene_var <- apply(neuronal_block_matrix, 1, var)

top_variable_genes <- names(sort(variable_gene_var, decreasing = TRUE))[1:min(2000, length(variable_gene_var))]

neuronal_block_pca <- prcomp(t(neuronal_block_matrix[top_variable_genes, ]), scale. = TRUE)

pca_variance_explained <- tibble(
  pc = factor(paste0("PC", seq_along(neuronal_block_pca$sdev)), levels = paste0("PC", seq_along(neuronal_block_pca$sdev))),
  variance_explained = neuronal_block_pca$sdev^2 / sum(neuronal_block_pca$sdev^2),
  cumulative_variance = cumsum(neuronal_block_pca$sdev^2) / sum(neuronal_block_pca$sdev^2)
)

knitr::kable(pca_variance_explained, digits = 3)

ggplot(pca_variance_explained, aes(x = pc, y = cumulative_variance, group = 1)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "grey40") +
  labs(
    title = "Cumulative variance explained across collinear neuronal subtypes",
    subtitle = "Dashed line = 90% variance; the PC count crossing it is the effective linear rank of this block",
    x = NULL, y = "Cumulative variance explained"
  ) +
  theme_classic()

nonneg_block_matrix <- pmax(neuronal_block_matrix[top_variable_genes, ] - min(neuronal_block_matrix[top_variable_genes, ]), 1e-6)

nmf_ranks_to_test <- 2:min(length(collinear_neuronal_subtypes), 6)

nmf_fit_quality <- map_dfr(nmf_ranks_to_test, function(k) {
  fit <- nmf(nonneg_block_matrix, rank = k, seed = 197289, .options = "v0")
  tibble(rank = k, residual_rss = sum((nonneg_block_matrix - fitted(fit))^2))
})

knitr::kable(nmf_fit_quality, digits = 1, caption = "NMF residual sum-of-squares by rank (elbow = effective non-negative rank)")

write.csv(
  pca_variance_explained,
  "results/marker_refinement/neuronal_block_pca_variance.csv",
  row.names = FALSE
)

write.csv(
  nmf_fit_quality,
  "results/marker_refinement/neuronal_block_nmf_rank.csv",
  row.names = FALSE
)
```

## Expanded marker manifest and re-running VIF diagnostics

```{r}
#| label: expanded-marker-manifest

expanded_neuronal_markers <- one_vs_rest_top20 |>
  transmute(gene, marker_group = subtype_author)

non_neuronal_markers_original <- combined_marker_manifest_original |>
  filter(!marker_group %in% unique(tg_neurons_sct$subtype_author))

expanded_marker_manifest <- bind_rows(
  expanded_neuronal_markers |> select(gene, marker_group),
  non_neuronal_markers_original |> select(gene, marker_group)
) |>
  distinct(gene, marker_group, .keep_all = TRUE) |>
  filter(gene %in% rownames(combined_logcpm))

manifest_comparison <- tibble(
  metric = c("Original manifest size", "Expanded manifest size", "Net new neuronal markers added"),
  value = c(
    nrow(combined_marker_manifest_original),
    nrow(expanded_marker_manifest),
    nrow(expanded_neuronal_markers |> anti_join(combined_marker_manifest_original, by = c("gene", "marker_group")))
  )
)

knitr::kable(manifest_comparison)

write.csv(
  expanded_marker_manifest,
  "results/marker_refinement/GSE197289_TG_expanded_marker_manifest.csv",
  row.names = FALSE
)
```

```{r}
#| label: fig-refit-vif-comparison
#| fig-width: 10
#| fig-height: 6

rebuild_signature_matrix <- function(manifest, logcpm_matrix, metadata) {
  genes <- manifest |> filter(gene %in% rownames(logcpm_matrix)) |> pull(gene) |> unique()
  linear_expr <- 2^logcpm_matrix[genes, , drop = FALSE]
  id_map <- metadata |>
    transmute(pseudobulk_id, reference_label = as.character(reference_label)) |>
    filter(pseudobulk_id %in% colnames(linear_expr))
  identity_levels <- unique(id_map$reference_label)
  sig <- sapply(identity_levels, function(id_lab) {
    cols <- id_map$pseudobulk_id[id_map$reference_label == id_lab]
    if (length(cols) == 1) linear_expr[, cols] else rowMeans(linear_expr[, cols, drop = FALSE])
  })
  colnames(sig) <- identity_levels
  sig
}

compute_vif <- function(x) {
  vif_out <- map_dbl(seq_len(ncol(x)), function(i) {
    y <- x[, i]
    preds <- x[, -i, drop = FALSE]
    fit <- lm(y ~ preds)
    r2 <- summary(fit)$r.squared
    if (r2 >= 0.999999) return(Inf)
    1 / (1 - r2)
  })
  names(vif_out) <- colnames(x)
  vif_out
}

expanded_signature_matrix <- rebuild_signature_matrix(expanded_marker_manifest, combined_logcpm, combined_metadata)
expanded_log_signature <- log2(expanded_signature_matrix + 1)

expanded_vif <- compute_vif(expanded_log_signature)
expanded_condition_number <- kappa(expanded_log_signature, exact = TRUE)

vif_before_after <- baseline_vif_table |>
  select(reference_identity, vif_before = vif) |>
  inner_join(
    tibble(reference_identity = names(expanded_vif), vif_after = expanded_vif),
    by = "reference_identity"
  ) |>
  arrange(vif_before)

knitr::kable(vif_before_after, digits = 2)

message(
  "Condition number before (combined manifest): ", round(kappa(log2(expanded_signature_matrix + 1), exact = TRUE), 1),
  " | after (expanded manifest): ", round(expanded_condition_number, 1)
)

vif_before_after_long <- vif_before_after |>
  pivot_longer(c(vif_before, vif_after), names_to = "stage", values_to = "vif") |>
  mutate(stage = recode(stage, vif_before = "Original manifest", vif_after = "Expanded manifest"))

ggplot(vif_before_after_long, aes(x = reference_identity, y = vif, fill = stage)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "firebrick") +
  labs(
    title = "VIF before vs. after empirical marker expansion",
    subtitle = "Dashed line = severe-collinearity threshold (VIF = 10)",
    x = NULL, y = "Variance inflation factor", fill = NULL
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

write.csv(
  vif_before_after,
  "results/marker_refinement/vif_before_after_expansion.csv",
  row.names = FALSE
)
```

## Summary and verdict

```{r}
#| label: refinement-summary

refinement_summary <- tibble(
  metric = c(
    "Mean silhouette width, collinear neuronal subtypes",
    "Mean VIF before expansion",
    "Mean VIF after expansion",
    "Net new empirically-specific markers added",
    "PCs needed for 90% variance in collinear block"
  ),
  value = c(
    round(mean(silhouette_summary$mean_silhouette[silhouette_summary$subtype_author %in% collinear_neuronal_subtypes]), 3),
    round(mean(vif_before_after$vif_before), 2),
    round(mean(vif_before_after$vif_after), 2),
    nrow(expanded_neuronal_markers |> anti_join(combined_marker_manifest_original, by = c("gene", "marker_group"))),
    min(which(pca_variance_explained$cumulative_variance >= 0.9))
  )
)

knitr::kable(refinement_summary)
```
