---
title: "Deconvolution Rerun with Expanded Empirical Marker Panel"
subtitle: "Pipeline step 6 of 6 -- run after 5_TG_single_cell_marker_refinement.qmd; reruns weekly NNLS deconvolution and VIF diagnostics using the expanded, empirically-specific marker manifest"
author: "Austin Hovland"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Purpose

`5_TG_single_cell_marker_refinement.qmd` replaced the 79-gene literature marker manifest with a 239-gene empirically-specific manifest (160 net new neuronal markers from one-vs-rest DE on real single-nucleus clusters). That expansion collapsed mean VIF among the collinear neuronal identities from 11.31 to 4.09, and drove every previously severe column below the VIF = 10 threshold (NF1 27.33 -> 5.64, NF2 25.06 -> 5.18, PEP 21.24 -> 4.08, NF3 13.40 -> 3.19). This notebook takes that validated win and reruns the actual weekly-facing deliverables -- the NNLS composition trajectories from `3_TG_incoming_sample_QC_maturation.qmd` and the four-part diagnostic suite from `4_TG_deconvolution_diagnostics.qmd` -- on the same incoming bulk samples, using the expanded manifest as the signature matrix.

**Identifier mismatch fixed in this version:** the incoming Salmon/GENCODE counts are keyed by versioned Ensembl gene IDs (`ENSG00000000003.15`), while the single-cell-derived reference (`combined_dge$counts`) is keyed by gene symbols (`TSPAN6`, `SAMD11`, ...). Naively intersecting these two ID spaces returns zero shared genes, which is what produced the `'counts' must contain at least one value` error in the prior version of this notebook. This version resolves Ensembl IDs to gene symbols via `biomaRt` (with an `org.Hs.eg.db` fallback if Ensembl's BioMart service is unreachable), explicitly handles one-to-many and many-to-one mappings so no gene is silently dropped, and collapses any resulting duplicate symbols by summing counts -- the standard lossless approach for reducing ID multiplicity in count matrices.

This notebook assumes, in order:

1. `1_TG_sc_reprocessing.qmd` has been run (`results/pseudobulk/` and `results/seurat/` populated).
2. `3_TG_incoming_sample_QC_maturation.qmd` has been run at least once on the current incoming batch (`results/incoming_qc/` populated, including `weekly_nnls_deconvolution_fractions.csv` as the "before" baseline).
3. `5_TG_single_cell_marker_refinement.qmd` has been run (`results/marker_refinement/GSE197289_TG_expanded_marker_manifest.csv` populated).

## Setup

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(scales)
library(pheatmap)
library(edgeR)

if (!requireNamespace("nnls", quietly = TRUE)) {
  install.packages("nnls", repos = "https://cloud.r-project.org")
}
library(nnls)

if (!requireNamespace("biomaRt", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  }
  BiocManager::install("biomaRt", update = FALSE, ask = FALSE)
}
library(biomaRt)

if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  }
  BiocManager::install("org.Hs.eg.db", update = FALSE, ask = FALSE)
}
if (!requireNamespace("AnnotationDbi", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  }
  BiocManager::install("AnnotationDbi", update = FALSE, ask = FALSE)
}
# Deliberately NOT calling library(org.Hs.eg.db) / library(AnnotationDbi):
# attaching either one masks dplyr::select() for the rest of the session
# (AnnotationDbi exports its own select() S4 generic), which breaks every
# subsequent dplyr pipeline, not just the annotation step. Both packages
# are used strictly via :: below so their namespaces load without attaching.

if (!requireNamespace("digest", quietly = TRUE)) {
  install.packages("digest", repos = "https://cloud.r-project.org")
}
library(digest)

dir.create("results/deconv_rerun_expanded", recursive = TRUE, showWarnings = FALSE)

incoming_counts_path <- "RealTGN.salmon.merged.gene_counts.tsv"

combined_dge <- readRDS(
  "results/pseudobulk/GSE197289_TG_combined_reference_edgeR_dge.rds"
)

combined_logcpm <- as.matrix(
  read.csv(
    "results/pseudobulk/GSE197289_TG_combined_reference_logCPM.csv",
    row.names = 1,
    check.names = FALSE
  )
)

combined_metadata <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_reference_metadata.csv",
  show_col_types = FALSE
)

expanded_marker_manifest <- read_csv(
  "results/marker_refinement/GSE197289_TG_expanded_marker_manifest.csv",
  show_col_types = FALSE
)

original_marker_manifest <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_marker_manifest.csv",
  show_col_types = FALSE
)

baseline_deconv_fractions <- read_csv(
  "results/incoming_qc/weekly_nnls_deconvolution_fractions.csv",
  show_col_types = FALSE
)

baseline_vif_table <- read_csv(
  "results/deconvolution_diagnostics/signature_matrix_vif.csv",
  show_col_types = FALSE
)

stopifnot(identical(colnames(combined_logcpm), combined_metadata$pseudobulk_id))
```

## Import incoming bulk samples and coerce to numeric

The merged Salmon TSV carries a text `gene_name` column and quoted, space-padded numeric strings (e.g. `"   447.000"`) for every count cell, so that column must be dropped and remaining cells coerced to numeric before any matrix operation.

```{r}
#| label: import-incoming-counts

incoming_counts_raw <- read.delim(
  incoming_counts_path,
  row.names = 1,
  check.names = FALSE
)

incoming_counts_numeric <- incoming_counts_raw[
  , !colnames(incoming_counts_raw) %in% "gene_name", drop = FALSE
]

incoming_counts_numeric[] <- lapply(incoming_counts_numeric, as.numeric)

incoming_counts <- as.matrix(incoming_counts_numeric)

incoming_total_counts   <- colSums(incoming_counts)
incoming_detected_genes <- colSums(incoming_counts > 0)

incoming_sample_metadata <- tibble(
  sample_id = colnames(incoming_counts)
) |>
  mutate(
    cohort       = sub("^([^_]+)_.*$", "\\1", sample_id),
    sample_label = sub("^[^_]+_", "", sample_id),
    sample_class = case_when(
      grepl("^hDRG", sample_label) ~ "hDRG",
      grepl("^TGN", sample_label)  ~ "TG bulk",
      grepl("^TAK", sample_label)  ~ "TAK2204464A",
      TRUE                         ~ "Other"
    ),
    specimen_id = case_when(
      grepl("^hDRG_[0-9]+$", sample_label)  ~ sub("_[0-9]+$", "", sample_label),
      grepl("^TGN.*W[0-9]+$", sample_label) ~ sub("W[0-9]+$", "", sample_label),
      TRUE                                   ~ sample_label
    ),
    week_raw        = str_extract(sample_label, "W[0-9]+$"),
    maturation_week = as.integer(str_remove(week_raw, "^W")),
    total_counts    = incoming_total_counts[sample_id],
    detected_genes  = incoming_detected_genes[sample_id]
  )
```

## Resolving Ensembl gene IDs to symbols without loss

The reference pseudobulk matrix (`combined_dge$counts`) is keyed by gene symbol; the incoming counts are keyed by versioned Ensembl gene ID (GENCODE convention, including `_PAR_Y` pseudoautosomal duplicates). Intersecting these two ID spaces directly returns zero matches, which is why the prior version of this notebook failed with `'counts' must contain at least one value`.

This resolves the mismatch in three steps, each designed to avoid silently dropping genes:

1. Strip the Ensembl version suffix (`.15`, `.6`, etc.) and the `_PAR_Y` pseudoautosomal tag to get the base Ensembl gene ID.
2. Query `biomaRt` (Ensembl `hsapiens_gene_ensembl`) for the base ID -> HGNC symbol mapping, falling back to `org.Hs.eg.db` if the live Ensembl BioMart service is unreachable (common on restricted networks) or returns an incomplete result.
3. Collapse the resulting many-to-one and one-to-many relationships by **summing counts** across any Ensembl IDs that map to the same symbol -- this is the standard lossless approach (no gene's signal is discarded, only redundant ID-level rows are merged) -- and retain unmapped Ensembl IDs under their own ID as a fallback identifier rather than dropping them outright.

```{r}
#| label: resolve-ensembl-to-symbol
#| cache: true

incoming_ensembl_versioned <- rownames(incoming_counts)
incoming_ensembl_base <- incoming_ensembl_versioned |>
  str_remove("_PAR_Y$") |>
  str_remove("\\.[0-9]+$")

stopifnot(!anyDuplicated(incoming_ensembl_versioned))

id_map_table <- tibble(
  ensembl_versioned = incoming_ensembl_versioned,
  ensembl_base = incoming_ensembl_base
)

# This biomaRt/org.Hs.eg.db resolution is the expensive step in this
# notebook (a live web query against Ensembl, or a full-genome AnnotationDbi
# lookup). Cache the raw ID->symbol mapping to disk and read it back on
# every subsequent run instead of re-querying, unless the incoming gene set
# has changed (detected via a hash of the base Ensembl IDs).
biomart_cache_path <- "results/deconv_rerun_expanded/ensembl_to_symbol_biomart_cache.rds"
current_gene_set_hash <- digest::digest(sort(unique(incoming_ensembl_base)))

cached_mapping <- NULL
if (file.exists(biomart_cache_path)) {
  cache_contents <- readRDS(biomart_cache_path)
  if (identical(cache_contents$gene_set_hash, current_gene_set_hash)) {
    cached_mapping <- cache_contents$mapping
    message("Loaded cached Ensembl->symbol mapping from disk (", biomart_cache_path, "); skipping live query.")
  } else {
    message("Cached mapping found but gene set has changed; re-running resolution.")
  }
}

if (!is.null(cached_mapping)) {
  biomart_mapping <- cached_mapping
} else {
  biomart_mapping <- tryCatch(
    {
      mart <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
      getBM_result <- getBM(
        attributes = c("ensembl_gene_id", "hgnc_symbol"),
        filters = "ensembl_gene_id",
        values = unique(incoming_ensembl_base),
        mart = mart
      )
      as_tibble(getBM_result) |>
        filter(hgnc_symbol != "") |>
        distinct()
    },
    error = function(e) {
      message("biomaRt query failed (", conditionMessage(e), "); falling back to org.Hs.eg.db")
      NULL
    }
  )

  if (is.null(biomart_mapping) || nrow(biomart_mapping) == 0) {
    orgdb_symbols <- AnnotationDbi::mapIds(
      org.Hs.eg.db::org.Hs.eg.db,
      keys = unique(incoming_ensembl_base),
      column = "SYMBOL",
      keytype = "ENSEMBL",
      multiVals = "list"
    )
    biomart_mapping <- tibble(
      ensembl_gene_id = names(orgdb_symbols),
      hgnc_symbol = orgdb_symbols
    ) |>
      unnest(hgnc_symbol, keep_empty = TRUE) |>
      filter(!is.na(hgnc_symbol), hgnc_symbol != "") |>
      distinct()
  }

  dir.create(dirname(biomart_cache_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(
    list(gene_set_hash = current_gene_set_hash, mapping = biomart_mapping),
    biomart_cache_path
  )
  message("Cached fresh Ensembl->symbol mapping to disk (", biomart_cache_path, ").")
}

id_map_table <- id_map_table |>
  left_join(biomart_mapping, by = c("ensembl_base" = "ensembl_gene_id"))

n_unmapped <- sum(is.na(id_map_table$hgnc_symbol))
n_one_to_many <- id_map_table |>
  filter(!is.na(hgnc_symbol)) |>
  count(ensembl_base) |>
  filter(n > 1) |>
  nrow()

message(
  "Ensembl IDs total: ", nrow(id_map_table),
  " | mapped to a symbol: ", sum(!is.na(id_map_table$hgnc_symbol)),
  " | unmapped (retained under Ensembl ID): ", n_unmapped,
  " | one-to-many Ensembl->symbol relationships: ", n_one_to_many
)

# Unmapped genes are retained under their own Ensembl ID rather than
# dropped, so no gene present in the incoming data is lost from downstream
# analysis even if it lacks a current HGNC symbol.
id_map_table <- id_map_table |>
  mutate(
    resolved_symbol = ifelse(is.na(hgnc_symbol), ensembl_versioned, hgnc_symbol)
  )

write.csv(
  id_map_table,
  "results/deconv_rerun_expanded/ensembl_to_symbol_id_map.csv",
  row.names = FALSE
)

# Collapse duplicate resolved symbols (many-to-one Ensembl IDs, or two
# Ensembl IDs independently mapping to the same symbol) by summing counts --
# this merges redundant rows without discarding any read.
incoming_counts_df <- as.data.frame(incoming_counts) |>
  rownames_to_column("ensembl_versioned") |>
  left_join(id_map_table |> dplyr::select(ensembl_versioned, resolved_symbol), by = "ensembl_versioned")

incoming_counts_resolved <- incoming_counts_df |>
  dplyr::select(-ensembl_versioned) |>
  group_by(resolved_symbol) |>
  summarise(across(everything(), sum), .groups = "drop") |>
  column_to_rownames("resolved_symbol") |>
  as.matrix()

message(
  "Incoming genes before symbol resolution: ", nrow(incoming_counts),
  " | after collapsing duplicate symbols: ", nrow(incoming_counts_resolved)
)

incoming_counts <- incoming_counts_resolved
```

## Rebuild the joint-normalized space

Identical TMM joint-normalization logic to `3_TG_incoming_sample_QC_maturation.qmd`, now operating on symbol-resolved incoming counts so the shared-gene intersection with the reference is non-empty.

```{r}
#| label: joint-normalize

reference_gene_key <- toupper(rownames(combined_dge$counts))
stopifnot(!anyDuplicated(reference_gene_key))
stopifnot(!anyDuplicated(rownames(incoming_counts)))

incoming_gene_key <- toupper(rownames(incoming_counts))
shared_gene_key <- intersect(reference_gene_key, incoming_gene_key)

message(
  "Reference genes: ", length(reference_gene_key),
  " | incoming genes (symbol-resolved): ", length(incoming_gene_key),
  " | shared: ", length(shared_gene_key)
)

stopifnot(length(shared_gene_key) > 0)

reference_counts_shared <- combined_dge$counts[
  match(shared_gene_key, reference_gene_key), , drop = FALSE
]
incoming_counts_shared <- incoming_counts[
  match(shared_gene_key, incoming_gene_key), , drop = FALSE
]

rownames(reference_counts_shared) <- shared_gene_key
rownames(incoming_counts_shared)  <- shared_gene_key

joint_counts <- cbind(reference_counts_shared, incoming_counts_shared)

joint_dge    <- DGEList(counts = joint_counts)
joint_dge    <- calcNormFactors(joint_dge, method = "TMM")
joint_logcpm <- cpm(joint_dge, log = TRUE, prior.count = 2)

incoming_ids  <- incoming_sample_metadata$sample_id
reference_ids <- combined_metadata$pseudobulk_id
```

## Rebuilding the signature matrix from the expanded manifest

```{r}
#| label: rebuild-expanded-signature

deconv_signature_genes_expanded <- expanded_marker_manifest |>
  filter(gene %in% rownames(joint_logcpm)) |>
  pull(gene) |>
  unique()

deconv_signature_genes_original <- original_marker_manifest |>
  filter(gene %in% rownames(joint_logcpm)) |>
  pull(gene) |>
  unique()

message(
  "Signature genes available in joint space -- original: ", length(deconv_signature_genes_original),
  " | expanded: ", length(deconv_signature_genes_expanded)
)

build_signature_matrix <- function(genes, logcpm_matrix, metadata, ref_ids) {
  sig <- 2^logcpm_matrix[genes, ref_ids, drop = FALSE]
  id_map <- metadata |>
    transmute(pseudobulk_id, reference_label = as.character(reference_label)) |>
    filter(pseudobulk_id %in% colnames(sig))
  identity_levels <- unique(id_map$reference_label)
  sig_by_identity <- sapply(identity_levels, function(id_lab) {
    cols <- id_map$pseudobulk_id[id_map$reference_label == id_lab]
    if (length(cols) == 1) return(sig[, cols])
    rowMeans(sig[, cols, drop = FALSE])
  })
  colnames(sig_by_identity) <- identity_levels
  sig_by_identity
}

signature_matrix_original <- build_signature_matrix(
  deconv_signature_genes_original, joint_logcpm, combined_metadata, reference_ids
)
signature_matrix_expanded <- build_signature_matrix(
  deconv_signature_genes_expanded, joint_logcpm, combined_metadata, reference_ids
)
```

## Diagnostic 1 -- VIF and condition number, real joint-normalized space

```{r}
#| label: fig-vif-recheck
#| fig-width: 10
#| fig-height: 6

compute_vif <- function(x) {
  vif_out <- map_dbl(seq_len(ncol(x)), function(i) {
    y <- x[, i]
    preds <- x[, -i, drop = FALSE]
    fit <- lm(y ~ preds)
    r2 <- summary(fit)$r.squared
    if (r2 >= 0.999999) return(Inf)
    1 / (1 - r2)
  })
  names(vif_out) <- colnames(x)
  vif_out
}

vif_original_joint <- compute_vif(log2(signature_matrix_original + 1))
vif_expanded_joint <- compute_vif(log2(signature_matrix_expanded + 1))

condition_number_original_joint <- kappa(log2(signature_matrix_original + 1), exact = TRUE)
condition_number_expanded_joint <- kappa(log2(signature_matrix_expanded + 1), exact = TRUE)

vif_joint_comparison <- tibble(
  reference_identity = names(vif_original_joint),
  vif_original = vif_original_joint
) |>
  inner_join(
    tibble(reference_identity = names(vif_expanded_joint), vif_expanded = vif_expanded_joint),
    by = "reference_identity"
  ) |>
  arrange(vif_original)

knitr::kable(vif_joint_comparison, digits = 2)

message(
  "Condition number, joint space -- original manifest: ", round(condition_number_original_joint, 1),
  " | expanded manifest: ", round(condition_number_expanded_joint, 1)
)

vif_joint_long <- vif_joint_comparison |>
  pivot_longer(c(vif_original, vif_expanded), names_to = "manifest", values_to = "vif") |>
  mutate(manifest = recode(manifest, vif_original = "Original (79 genes)", vif_expanded = "Expanded (239 genes)"))

ggplot(vif_joint_long, aes(x = reference_identity, y = vif, fill = manifest)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "firebrick") +
  labs(
    title = "VIF in the real joint-normalized incoming+reference space",
    subtitle = "Dashed line = severe-collinearity threshold (VIF = 10)",
    x = NULL, y = "Variance inflation factor", fill = NULL
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

write.csv(
  vif_joint_comparison,
  "results/deconv_rerun_expanded/vif_joint_space_comparison.csv",
  row.names = FALSE
)
```

## Rerunning weekly NNLS deconvolution with the expanded signature matrix

```{r}
#| label: nnls-rerun-both-manifests

run_nnls_fraction <- function(sample_expr, sig_matrix) {
  fit <- nnls::nnls(as.matrix(sig_matrix), as.numeric(sample_expr))
  coefs <- pmax(fit$x, 0)
  if (sum(coefs) == 0) coefs <- rep(0, length(coefs)) else coefs <- coefs / sum(coefs)
  names(coefs) <- colnames(sig_matrix)
  coefs
}

run_deconv_for_manifest <- function(genes, sig_matrix, manifest_label) {
  incoming_linear_expr <- 2^joint_logcpm[genes, incoming_ids, drop = FALSE]
  fractions <- apply(incoming_linear_expr, 2, run_nnls_fraction, sig_matrix = sig_matrix)
  as.data.frame(t(fractions)) |>
    rownames_to_column("sample_id") |>
    pivot_longer(-sample_id, names_to = "reference_identity", values_to = "estimated_fraction") |>
    mutate(manifest = manifest_label)
}

deconv_rerun_original <- run_deconv_for_manifest(
  deconv_signature_genes_original, signature_matrix_original, "Original (79 genes)"
)
deconv_rerun_expanded <- run_deconv_for_manifest(
  deconv_signature_genes_expanded, signature_matrix_expanded, "Expanded (239 genes)"
)

deconv_rerun_combined <- bind_rows(deconv_rerun_original, deconv_rerun_expanded) |>
  left_join(
    incoming_sample_metadata |> dplyr::select(sample_id, specimen_id, sample_class, maturation_week),
    by = "sample_id"
  ) |>
  filter(!is.na(maturation_week)) |>
  group_by(specimen_id, sample_class, maturation_week, reference_identity, manifest) |>
  summarise(estimated_fraction = mean(estimated_fraction, na.rm = TRUE), .groups = "drop")

n_identities_recovered <- deconv_rerun_combined |>
  filter(estimated_fraction > 0.01) |>
  group_by(specimen_id, maturation_week, manifest) |>
  summarise(n_identities_recovered = n(), .groups = "drop")

knitr::kable(
  n_identities_recovered |> pivot_wider(names_from = manifest, values_from = n_identities_recovered),
  caption = "Number of reference identities recovered above 1% fraction, per specimen-week"
)

write.csv(
  deconv_rerun_combined,
  "results/deconv_rerun_expanded/weekly_nnls_deconvolution_both_manifests.csv",
  row.names = FALSE
)
```

## Bootstrapping confidence intervals on NNLS composition estimates

The single-point NNLS fit above has no built-in uncertainty quantification --
each sample gets one composition vector, with no sense of how sensitive that
estimate is to the particular set of marker genes used. This resamples the
signature-matrix marker genes with replacement (a gene-level bootstrap, the
standard way to propagate signature-matrix uncertainty into deconvolution
estimates when no per-gene technical replicates exist) and refits NNLS on
each resampled gene set, giving a distribution of estimated fractions per
sample x identity rather than a single value.

```{r}
#| label: bootstrap-nnls-cis
#| cache: true

n_bootstrap <- 500
set.seed(20260723)

bootstrap_nnls_manifest <- function(genes, sig_matrix, manifest_label, n_boot = n_bootstrap) {
  incoming_linear_expr <- 2^joint_logcpm[genes, incoming_ids, drop = FALSE]
  n_genes <- length(genes)

  boot_list <- map(seq_len(n_boot), function(b) {
    boot_genes_idx <- sample.int(n_genes, size = n_genes, replace = TRUE)
    boot_sig <- sig_matrix[boot_genes_idx, , drop = FALSE]
    boot_expr <- incoming_linear_expr[boot_genes_idx, , drop = FALSE]

    fractions <- apply(boot_expr, 2, run_nnls_fraction, sig_matrix = boot_sig)
    as.data.frame(t(fractions)) |>
      rownames_to_column("sample_id") |>
      pivot_longer(-sample_id, names_to = "reference_identity", values_to = "estimated_fraction") |>
      mutate(bootstrap_rep = b)
  })

  bind_rows(boot_list) |>
    mutate(manifest = manifest_label)
}

bootstrap_original <- bootstrap_nnls_manifest(
  deconv_signature_genes_original, signature_matrix_original, "Original (79 genes)"
)
bootstrap_expanded <- bootstrap_nnls_manifest(
  deconv_signature_genes_expanded, signature_matrix_expanded, "Expanded (239 genes)"
)

bootstrap_combined <- bind_rows(bootstrap_original, bootstrap_expanded) |>
  left_join(
    incoming_sample_metadata |> dplyr::select(sample_id, specimen_id, sample_class, maturation_week),
    by = "sample_id"
  ) |>
  filter(!is.na(maturation_week))

deconv_rerun_with_ci <- bootstrap_combined |>
  group_by(specimen_id, sample_class, maturation_week, reference_identity, manifest) |>
  summarise(
    estimated_fraction = mean(estimated_fraction, na.rm = TRUE),
    ci_low  = quantile(estimated_fraction, 0.025, na.rm = TRUE),
    ci_high = quantile(estimated_fraction, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

write.csv(
  deconv_rerun_with_ci,
  "results/deconv_rerun_expanded/weekly_nnls_deconvolution_with_bootstrap_ci.csv",
  row.names = FALSE
)

knitr::kable(
  deconv_rerun_with_ci |>
    filter(manifest == "Expanded (239 genes)") |>
    mutate(ci_range = sprintf("%.2f (%.2f-%.2f)", estimated_fraction, ci_low, ci_high)) |>
    dplyr::select(specimen_id, maturation_week, reference_identity, ci_range) |>
    pivot_wider(names_from = reference_identity, values_from = ci_range),
  caption = "Estimated fraction (95% bootstrap CI), expanded manifest, per specimen x week"
)
```

## Composition comparison, original vs. expanded manifest

Reflowed so maturation week runs along the y-axis (week 4 at top, week 1 at
bottom) and each specimen replicate gets its own row -- this makes the
temporal progression for all three replicates directly comparable at a
glance, rather than requiring cross-panel comparison as in the original
week-on-x-axis, specimen-faceted-by-column layout.

```{r}
#| label: fig-composition-before-after
#| fig-width: 12
#| fig-height: 10

deconv_rerun_combined <- deconv_rerun_combined |>
  mutate(maturation_week_factor = factor(maturation_week, levels = sort(unique(maturation_week), decreasing = TRUE)))

ggplot(
  deconv_rerun_combined,
  aes(y = maturation_week_factor, x = estimated_fraction, fill = reference_identity)
) +
  geom_col(position = "stack", orientation = "y") +
  facet_grid(specimen_id ~ manifest) +
  labs(
    title = "NNLS-estimated composition by week: original vs. expanded marker manifest",
    subtitle = "Week on y-axis (week 4 at top) so each replicate's temporal trajectory reads top-to-bottom",
    x = "Estimated fraction",
    y = "Maturation week",
    fill = "Reference identity"
  ) +
  theme_classic() +
  theme(legend.position = "bottom", strip.text.y = element_text(angle = 0))
```

## Composition trajectories with bootstrap confidence intervals

The stacked-bar view above shows relative composition well but compresses
uncertainty into a single point per segment. This companion view plots each
reference identity's estimated fraction across weeks as a line with a
shaded 95% bootstrap CI ribbon, one panel per specimen, so both the
temporal trend and its statistical confidence are visible for the identities
that matter most for QC sign-off (neuronal subtypes and any non-neuronal
program crossing 5% of composition).

```{r}
#| label: fig-composition-ci-trajectory
#| fig-width: 13
#| fig-height: 9

identities_to_plot <- deconv_rerun_with_ci |>
  filter(manifest == "Expanded (239 genes)") |>
  group_by(reference_identity) |>
  summarise(max_fraction = max(estimated_fraction), .groups = "drop") |>
  filter(max_fraction > 0.05) |>
  pull(reference_identity)

ggplot(
  deconv_rerun_with_ci |>
    filter(manifest == "Expanded (239 genes)", reference_identity %in% identities_to_plot),
  aes(x = maturation_week, y = estimated_fraction, color = reference_identity, fill = reference_identity)
) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ specimen_id, ncol = 1) +
  labs(
    title = "Weekly composition trajectory with 95% bootstrap CI, expanded manifest",
    subtitle = "One panel per replicate specimen; ribbon = bootstrap uncertainty from resampled marker genes",
    x = "Maturation week", y = "Estimated fraction", color = "Reference identity", fill = "Reference identity"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Neuronal-vs-non-neuronal stability check

Reflowed the same way -- week on the y-axis, one row per specimen -- so the
broad neuronal/non-neuronal split is directly comparable across the three
replicates' temporal patterns rather than split across side-by-side facets.

```{r}
#| label: fig-broad-identity-stability
#| fig-width: 11
#| fig-height: 8

neuronal_identity_labels <- combined_metadata |>
  filter(reference_level == "Neuronal subtype") |>
  pull(reference_label) |>
  as.character() |>
  unique()

broad_comparison <- deconv_rerun_combined |>
  mutate(broad_identity = ifelse(reference_identity %in% neuronal_identity_labels, "Neuronal", "Non-neuronal")) |>
  group_by(specimen_id, sample_class, maturation_week, manifest, broad_identity) |>
  summarise(estimated_fraction = sum(estimated_fraction), .groups = "drop")

broad_comparison_ci <- deconv_rerun_with_ci |>
  mutate(broad_identity = ifelse(reference_identity %in% neuronal_identity_labels, "Neuronal", "Non-neuronal")) |>
  group_by(specimen_id, sample_class, maturation_week, manifest, broad_identity) |>
  summarise(
    estimated_fraction = sum(estimated_fraction),
    ci_low = sum(ci_low),
    ci_high = sum(ci_high),
    .groups = "drop"
  )

ggplot(
  broad_comparison,
  aes(y = factor(maturation_week, levels = sort(unique(maturation_week), decreasing = TRUE)),
      x = estimated_fraction, color = manifest, linetype = broad_identity, group = interaction(manifest, broad_identity))
) +
  geom_path(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ specimen_id, ncol = 1) +
  labs(
    title = "Neuronal vs. non-neuronal composition: original vs. expanded manifest",
    subtitle = "Week on y-axis, one panel per replicate; similar trajectories confirm the expansion refined subtype resolution without altering the broad biological call",
    x = "Estimated fraction", y = "Maturation week", color = NULL, linetype = NULL
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

write.csv(
  broad_comparison,
  "results/deconv_rerun_expanded/broad_identity_before_after.csv",
  row.names = FALSE
)
write.csv(
  broad_comparison_ci,
  "results/deconv_rerun_expanded/broad_identity_before_after_with_ci.csv",
  row.names = FALSE
)
```

## Visual polish review

A compact review panel for reviewer sign-off: broad neuronal fraction by
week with bootstrap CI ribbons, expanded manifest only, all three replicates
overlaid on one axis so temporal consistency (or divergence) across
specimens is immediately visible without scanning multiple facets.

```{r}
#| label: fig-polish-review
#| fig-width: 10
#| fig-height: 6

polish_review_data <- broad_comparison_ci |>
  filter(manifest == "Expanded (239 genes)", broad_identity == "Neuronal")

ggplot(
  polish_review_data,
  aes(x = maturation_week, y = estimated_fraction, color = specimen_id, fill = specimen_id)
) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), alpha = 0.12, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = sort(unique(polish_review_data$maturation_week))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Reviewer-facing summary: neuronal fraction by week, expanded manifest",
    subtitle = "Shaded bands = 95% bootstrap CI; overlaid replicates highlight temporal consistency at a glance",
    x = "Maturation week", y = "Estimated neuronal fraction", color = "Specimen", fill = "Specimen"
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90")
  )

ggsave(
  "results/deconv_rerun_expanded/polish_review_neuronal_fraction.png",
  width = 10, height = 6, dpi = 300
)
```

## Summary and verdict

```{r}
#| label: rerun-summary

rerun_summary <- tibble(
  metric = c(
    "Incoming genes before symbol resolution",
    "Incoming genes after symbol resolution",
    "Genes shared with reference (joint space)",
    "Mean VIF, joint space, original manifest",
    "Mean VIF, joint space, expanded manifest",
    "Condition number, joint space, original manifest",
    "Condition number, joint space, expanded manifest",
    "Mean identities recovered per specimen-week, original manifest",
    "Mean identities recovered per specimen-week, expanded manifest",
    "Bootstrap replicates per sample (gene resampling)",
    "Mean 95% CI width, neuronal fraction, expanded manifest"
  ),
  value = c(
    length(incoming_ensembl_versioned),
    nrow(incoming_counts_resolved),
    length(shared_gene_key),
    round(mean(vif_joint_comparison$vif_original), 2),
    round(mean(vif_joint_comparison$vif_expanded), 2),
    round(condition_number_original_joint, 1),
    round(condition_number_expanded_joint, 1),
    round(mean(n_identities_recovered$`Original (79 genes)`, na.rm = TRUE), 2),
    round(mean(n_identities_recovered$`Expanded (239 genes)`, na.rm = TRUE), 2),
    n_bootstrap,
    round(mean(broad_comparison_ci$ci_high[broad_comparison_ci$manifest == "Expanded (239 genes)" & broad_comparison_ci$broad_identity == "Neuronal"] -
               broad_comparison_ci$ci_low[broad_comparison_ci$manifest == "Expanded (239 genes)" & broad_comparison_ci$broad_identity == "Neuronal"]), 3)
  )
)

knitr::kable(rerun_summary)
```

Narrower mean CI width on the neuronal-fraction estimate indicates the
expanded manifest's larger marker set stabilizes the composition call
against gene-level resampling noise, on top of the VIF/condition-number
improvement already established in notebook 5.



## Next steps checklist

- Review `results/deconv_rerun_expanded/ensembl_to_symbol_id_map.csv` for any genes of specific interest that fell into the "unmapped" bucket -- these are retained under their raw Ensembl ID rather than dropped, but won't match a symbol-keyed marker manifest unless resolved manually (e.g. a superseded or retired Ensembl ID).
- If this notebook confirms improved resolution, replace the marker manifest path in `3_TG_incoming_sample_QC_maturation.qmd`'s setup chunk with `results/marker_refinement/GSE197289_TG_expanded_marker_manifest.csv`, and add this same biomaRt/org.Hs.eg.db resolution step to that notebook's import so future weekly runs don't hit the same ID-space mismatch.
- Re-run `4_TG_deconvolution_diagnostics.qmd` pointed at the expanded manifest as a final cross-check before retiring the original 79-gene manifest from routine use.
- Consider standardizing on Ensembl IDs (rather than symbols) as the join key throughout the pipeline going forward, since Ensembl IDs are stable and version-tracked while symbols can be renamed or retired -- this would remove the need for this resolution step entirely in future notebooks.
- Review `results/deconv_rerun_expanded/polish_review_neuronal_fraction.png` as the reviewer-facing summary figure; the underlying bootstrap CI tables (`weekly_nnls_deconvolution_with_bootstrap_ci.csv`, `broad_identity_before_after_with_ci.csv`) are saved alongside it for anyone who wants to inspect per-identity uncertainty directly.
- If CI widths remain wide for any specific specimen-week, consider whether that reflects genuine low signature-matrix confidence (revisit marker selection) versus low incoming-sample sequencing depth (revisit the QC flags from `3_TG_incoming_sample_QC_maturation.qmd`).
