---
title: "Incoming RNA-seq Sample QC & Weekly Maturation Tracking"
subtitle: "TG-reference-anchored review of new bulk RNA-seq batches (RealTGN.salmon)"
author: "Austin Hovland"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Purpose

This notebook is the recurring intake-QC and maturation-tracking companion to
`TG_sc_reprocessing-3.qmd` (single-nucleus reference build) and
`TG_reviewer_figures.qmd` (bulk-to-reference projection). It is meant to be
re-run every time a new batch of bulk RNA-seq quantifications lands, and does
two things:

1. **Incoming sample QC** — library size, complexity, duplication/composition
   sanity checks, and similarity to the integrated TG single-nucleus reference,
   flagging any sample that looks degraded, contaminated, or mislabeled before
   it enters downstream analysis.
2. **Weekly maturation investigation** — tracks the curated marker sets
   already derived for the TG reference (pan-neuronal, peptidergic/nociceptor,
   non-peptidergic, neurofilament/mechanosensory, cold/pruriceptor, and the
   non-neuronal purity programs) across the differentiation time course
   encoded in each sample's well/week label (e.g. `TGN...W1`, `W2`, ...), so
   drift or stalled maturation is caught early. Replicates are condensed to
   one value per specimen x week before any signal is computed. Three
   complementary, reference-anchored methods are then used together: (a) a
   correlation-weighted composite gene-signature score per program (learned
   from the pseudobulk reference, not uniform per-gene weighting), (b) a PCA
   projection of incoming samples into a *fixed* pseudobulk-reference PCA
   space, drawn as a weekly trajectory arrow, and (c) an NNLS deconvolution
   that estimates each sample's transcriptional composition against the
   pseudobulk reference as a signature matrix. Agreement across all three is
   treated as the strongest evidence of genuine maturation.

This notebook assumes `TG_sc_reprocessing-3.qmd` has already been run (so
`results/pseudobulk/` is populated) and that new bulk counts are available as
a Salmon merged gene-counts TSV, following the same layout as
`RealTGN.salmon.merged.gene_counts.tsv`.

## Setup

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(edgeR)

dir.create("results/incoming_qc", recursive = TRUE, showWarnings = FALSE)

# --- Point this at the newest merged Salmon gene-counts table each week ---
incoming_counts_path <- "RealTGN.salmon.merged.gene_counts.tsv"

# Reference artifacts produced by TG_sc_reprocessing-3.qmd
combined_dge <- readRDS(
  "results/pseudobulk/GSE197289_TG_combined_reference_edgeR_dge.rds"
)

combined_logcpm <- as.matrix(
  read.csv(
    "results/pseudobulk/GSE197289_TG_combined_reference_logCPM.csv",
    row.names = 1,
    check.names = FALSE
  )
)

combined_metadata <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_reference_metadata.csv",
  show_col_types = FALSE
)

# Previously derived marker manifest (pan-neuronal, subtype, non-neuronal programs)
combined_marker_manifest <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_marker_manifest.csv",
  show_col_types = FALSE
)

stopifnot(identical(colnames(combined_logcpm), combined_metadata$pseudobulk_id))
```

## Import incoming bulk RNA-seq samples

```{r}
#| label: import-incoming-samples

incoming_quant <- read_tsv(
  incoming_counts_path,
  show_col_types = FALSE
)

stopifnot(all(c("gene_id", "gene_name") %in% colnames(incoming_quant)))

incoming_sample_ids <- setdiff(colnames(incoming_quant), c("gene_id", "gene_name"))

stopifnot(length(incoming_sample_ids) > 0L)

incoming_counts_tbl <- incoming_quant |>
  transmute(
    gene_name = toupper(trimws(gene_name)),
    across(all_of(incoming_sample_ids), as.numeric)
  ) |>
  filter(!is.na(gene_name), gene_name != "") |>
  group_by(gene_name) |>
  summarise(
    across(all_of(incoming_sample_ids), \(x) sum(x, na.rm = TRUE)),
    .groups = "drop"
  )

incoming_counts <- incoming_counts_tbl |> as.data.frame()
rownames(incoming_counts) <- incoming_counts$gene_name
incoming_counts$gene_name <- NULL
incoming_counts <- as.matrix(incoming_counts)

stopifnot(!anyDuplicated(rownames(incoming_counts)))
```

## Parse sample identity and week/maturation timepoint

Sample labels follow the established convention: cohort prefix, then
`hDRG_<n>`, `TGN<label>W<week>`, or `TAK<lot>`. The trailing `W<number>`
token is parsed out explicitly as the differentiation week for the
maturation time course; samples without a parseable week are retained but
excluded from the weekly trajectory plots.

```{r}
#| label: parse-sample-identity

incoming_total_counts <- colSums(incoming_counts)
incoming_detected_genes <- colSums(incoming_counts > 0)

incoming_sample_metadata <- tibble(
  sample_id = colnames(incoming_counts)
) |>
  mutate(
    cohort = sub("^([^_]+)_.*$", "\\1", sample_id),
    sample_label = sub("^[^_]+_", "", sample_id),
    sample_class = case_when(
      grepl("^hDRG", sample_label) ~ "hDRG",
      grepl("^TGN", sample_label) ~ "TG bulk",
      grepl("^TAK", sample_label) ~ "TAK2204464A",
      TRUE ~ "Other"
    ),
    specimen_id = case_when(
      grepl("^hDRG_[0-9]+$", sample_label) ~ sub("_[0-9]+$", "", sample_label),
      grepl("^TGN.*W[0-9]+$", sample_label) ~ sub("W[0-9]+$", "", sample_label),
      TRUE ~ sample_label
    ),
    week_raw = str_extract(sample_label, "W[0-9]+$"),
    maturation_week = as.integer(str_remove(week_raw, "^W")),
    total_counts = incoming_total_counts[sample_id],
    detected_genes = incoming_detected_genes[sample_id],
    genes_per_million_counts = 1e6 * detected_genes / total_counts
  )

n_missing_week <- sum(is.na(incoming_sample_metadata$maturation_week))

message(
  "Samples parsed: ", nrow(incoming_sample_metadata),
  "\nSamples without a parseable maturation week: ", n_missing_week
)

knitr::kable(
  incoming_sample_metadata |>
    select(sample_id, cohort, sample_class, specimen_id, maturation_week,
           total_counts, detected_genes),
  digits = 0
)

write.csv(
  incoming_sample_metadata,
  "results/incoming_qc/incoming_sample_metadata.csv",
  row.names = FALSE
)
```

## Incoming sample QC — library size and complexity

**QC thresholds** below mirror the conservative single-nucleus flags already
established in `TG_sc_reprocessing-3.qmd` (`qc_min_counts`, `qc_min_features`)
but rescaled for bulk libraries; adjust `bulk_min_total_counts` and
`bulk_min_detected_genes` as the pipeline's bulk QC baseline is refined.

```{r}
#| label: incoming-qc-thresholds
#| fig-width: 11
#| fig-height: 5

bulk_min_total_counts <- 5e6
bulk_min_detected_genes <- 10000L

incoming_sample_metadata <- incoming_sample_metadata |>
  mutate(
    qc_low_depth = total_counts < bulk_min_total_counts,
    qc_low_complexity = detected_genes < bulk_min_detected_genes,
    qc_flag = qc_low_depth | qc_low_complexity,
    qc_status = factor(
      ifelse(qc_flag, "flagged", "retain"),
      levels = c("retain", "flagged")
    )
  )

knitr::kable(
  incoming_sample_metadata |>
    filter(qc_flag) |>
    select(sample_id, sample_class, total_counts, detected_genes,
           qc_low_depth, qc_low_complexity),
  digits = 0
)

ggplot(
  incoming_sample_metadata,
  aes(x = total_counts, y = detected_genes, color = sample_class, shape = qc_status)
) +
  geom_vline(xintercept = bulk_min_total_counts, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = bulk_min_detected_genes, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = sample_id), size = 2.6, max.overlaps = 20, show.legend = FALSE) +
  scale_x_log10() +
  labs(
    title = "Incoming bulk RNA-seq library depth vs. complexity",
    subtitle = "Dashed lines mark the current bulk QC floor",
    x = "Total counts (log scale)",
    y = "Detected genes",
    color = "Sample class",
    shape = "QC status"
  ) +
  theme_classic()

write.csv(
  incoming_sample_metadata,
  "results/incoming_qc/incoming_sample_qc_flags.csv",
  row.names = FALSE
)
```

## Joint normalization with the TG reference

Incoming samples are jointly TMM-normalized with the filtered combined TG
pseudobulk reference over their shared gene set, exactly as in
`TG_reviewer_figures.qmd`, so every weekly batch lands on the same expression
scale as the reference before any similarity or marker comparison is made.

```{r}
#| label: joint-normalization

reference_gene_key <- toupper(rownames(combined_dge$counts))
stopifnot(!anyDuplicated(reference_gene_key))
stopifnot(!anyDuplicated(rownames(incoming_counts)))

shared_gene_key <- intersect(reference_gene_key, rownames(incoming_counts))

reference_counts_shared <- combined_dge$counts[
  match(shared_gene_key, reference_gene_key), , drop = FALSE
]
incoming_counts_shared <- incoming_counts[
  match(shared_gene_key, rownames(incoming_counts)), , drop = FALSE
]

rownames(reference_counts_shared) <- shared_gene_key
rownames(incoming_counts_shared) <- shared_gene_key

stopifnot(identical(rownames(reference_counts_shared), rownames(incoming_counts_shared)))
stopifnot(nrow(reference_counts_shared) > 10000L)

joint_counts <- cbind(reference_counts_shared, incoming_counts_shared)

joint_sample_metadata <- bind_rows(
  combined_metadata |>
    transmute(
      sample_id = pseudobulk_id,
      sample_origin = "Reference pseudobulk",
      reference_label = as.character(reference_label),
      reference_level = as.character(reference_level),
      cohort = paste0("donor_", donor),
      maturation_week = NA_integer_
    ),
  incoming_sample_metadata |>
    transmute(
      sample_id,
      sample_origin = "Incoming bulk RNA-seq",
      reference_label = sample_class,
      reference_level = specimen_id,
      cohort,
      maturation_week
    )
)

joint_sample_metadata <- joint_sample_metadata[
  match(colnames(joint_counts), joint_sample_metadata$sample_id), , drop = FALSE
]
stopifnot(identical(colnames(joint_counts), joint_sample_metadata$sample_id))

joint_dge <- DGEList(counts = joint_counts, samples = as.data.frame(joint_sample_metadata))
joint_dge <- calcNormFactors(joint_dge, method = "TMM")
joint_logcpm <- cpm(joint_dge, log = TRUE, prior.count = 2)

n_top_var_genes <- 2000
gene_var <- apply(combined_logcpm, 1, var)
top_var_genes <- names(sort(gene_var, decreasing = TRUE))[seq_len(min(n_top_var_genes, length(gene_var)))]
projection_genes <- intersect(toupper(top_var_genes), rownames(joint_logcpm))
stopifnot(length(projection_genes) >= 500L)
```

## Incoming sample similarity to TG reference identities

Flags samples whose strongest correlation is not to the expected neuronal
program (e.g. contamination, mislabeling, or a maturation arrest that looks
non-neuronal transcriptionally).

```{r}
#| label: incoming-reference-correlation
#| fig-width: 12
#| fig-height: 10

incoming_ids <- incoming_sample_metadata$sample_id
reference_ids <- combined_metadata$pseudobulk_id

incoming_reference_cor <- cor(
  joint_logcpm[projection_genes, incoming_ids, drop = FALSE],
  joint_logcpm[projection_genes, reference_ids, drop = FALSE],
  method = "spearman"
)

incoming_reference_long <- as.data.frame(as.table(incoming_reference_cor)) |>
  rename(bulk_sample = Var1, reference_id = Var2, spearman_correlation = Freq) |>
  left_join(
    combined_metadata |>
      transmute(
        reference_id = pseudobulk_id,
        reference_label = as.character(reference_label),
        reference_level = as.character(reference_level)
      ),
    by = "reference_id"
  )

incoming_reference_summary <- incoming_reference_long |>
  group_by(bulk_sample, reference_label, reference_level) |>
  summarise(median_correlation = median(spearman_correlation), .groups = "drop") |>
  arrange(bulk_sample, desc(median_correlation))

incoming_top_matches <- incoming_reference_summary |>
  group_by(bulk_sample) |>
  slice_max(order_by = median_correlation, n = 3, with_ties = FALSE) |>
  ungroup()

knitr::kable(incoming_top_matches, digits = 3)

incoming_expect_neuron_flag <- incoming_top_matches |>
  group_by(bulk_sample) |>
  slice_max(order_by = median_correlation, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    sample_id = as.character(bulk_sample),
    top_reference_match = reference_label,
    top_reference_level = reference_level,
    top_match_correlation = median_correlation,
    identity_flag = !top_reference_level %in% c("Neuronal subtype")
  )

knitr::kable(
  incoming_expect_neuron_flag |> filter(identity_flag),
  digits = 3
)

write.csv(
  incoming_reference_summary,
  "results/incoming_qc/incoming_bulk_reference_correlation.csv",
  row.names = FALSE
)

write.csv(
  incoming_expect_neuron_flag,
  "results/incoming_qc/incoming_identity_flags.csv",
  row.names = FALSE
)
```

Reviewer-facing interpretation: the TG-bulk samples behave coherently
against the single-cell-derived reference, with their strongest matches
consistently falling among neuronal subtypes (especially NF2, NF3, and PEP),
while the external comparison samples (`hDRG`, `TAK2204464A`) correctly fail
that expectation and instead anchor to broad non-neuronal identities. That is
a strong sanity check on the reference projection logic rather than a QC
failure of the TG-bulk time course itself.

## Replicate-condensed weekly progression signal

When technical or biological replicates exist within the same specimen and
week, they should be condensed before downstream interpretation so the weekly
signal reflects progression rather than replicate noise. In the current batch,
all parsed TG-bulk specimen-week combinations are singletons (`n_replicates =
1` for each retained specimen-week), so this step acts mainly as a future-proof
aggregation layer rather than changing the current results.

```{r}
#| label: condense-replicates

maturation_markers_present <- combined_marker_manifest |>
  filter(gene %in% rownames(joint_logcpm))

incoming_logcpm_markers <- joint_logcpm[maturation_markers_present$gene, incoming_ids, drop = FALSE]

replicate_condensed_expr <- incoming_logcpm_markers |>
  as.data.frame() |>
  rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "sample_id", values_to = "log_cpm") |>
  left_join(maturation_markers_present, by = "gene") |>
  left_join(
    incoming_sample_metadata |> select(sample_id, specimen_id, sample_class, maturation_week),
    by = "sample_id"
  ) |>
  filter(!is.na(maturation_week)) |>
  group_by(specimen_id, sample_class, maturation_week, marker_group, gene) |>
  summarise(
    mean_log_cpm = mean(log_cpm, na.rm = TRUE),
    sd_log_cpm = sd(log_cpm, na.rm = TRUE),
    n_replicates = n(),
    .groups = "drop"
  )

replicate_summary_counts <- replicate_condensed_expr |>
  distinct(specimen_id, maturation_week, n_replicates) |>
  arrange(specimen_id, maturation_week)

knitr::kable(
  replicate_summary_counts,
  caption = "Replicate count per specimen x week going into the condensed signal"
)

write.csv(
  replicate_condensed_expr,
  "results/incoming_qc/replicate_condensed_marker_expression.csv",
  row.names = FALSE
)
```

In this run, the replicate-condensation layer is structurally correct but not
analytically active yet because each retained TG-bulk specimen-week appears
exactly once. That matters for interpretation: any week-to-week trend below is
currently a **specimen trajectory**, not a replicate-averaged consensus signal.

## Composite gene-signature score per program (replicate-condensed)

Rather than an unweighted per-gene z-score band, each marker program is
condensed to a single composite signature score per specimen-week using the
mean pairwise-correlation-weighted z-score across program genes (genes more
representative of the program, i.e. more correlated with the rest of the
program in the pseudobulk reference, contribute more). Weights are learned
from the pseudobulk reference only, then applied identically to the incoming
samples, so a stronger gene doesn't get diluted by a noisy or off-target one
within the same program.

```{r}
#| label: gene-signature-scoring

signature_weights <- map_dfr(unique(maturation_markers_present$marker_group), function(grp) {
  genes_in_group <- maturation_markers_present |>
    filter(marker_group == grp) |>
    pull(gene)

  genes_in_group <- intersect(genes_in_group, rownames(combined_logcpm))
  if (length(genes_in_group) < 3) {
    return(tibble(gene = genes_in_group, marker_group = grp, weight = 1))
  }

  ref_sub <- combined_logcpm[genes_in_group, , drop = FALSE]
  gene_cor <- cor(t(ref_sub), method = "spearman")
  mean_cor <- rowMeans(gene_cor) - (1 / nrow(gene_cor))
  weight <- pmax(mean_cor, 0.05)

  tibble(gene = genes_in_group, marker_group = grp, weight = weight / sum(weight) * length(weight))
})

incoming_full_z <- joint_logcpm[unique(signature_weights$gene), incoming_ids, drop = FALSE]
incoming_full_z <- t(scale(t(incoming_full_z)))
incoming_full_z <- incoming_full_z[apply(incoming_full_z, 1, function(x) all(is.finite(x))), , drop = FALSE]

reference_full_z <- joint_logcpm[rownames(incoming_full_z), reference_ids, drop = FALSE]
reference_full_z <- t(scale(t(reference_full_z)))
reference_full_z <- reference_full_z[apply(reference_full_z, 1, function(x) all(is.finite(x))), , drop = FALSE]

shared_sig_genes <- intersect(rownames(incoming_full_z), rownames(reference_full_z))
signature_weights <- signature_weights |> filter(gene %in% shared_sig_genes)

compute_signature_scores <- function(zmat, weights_df) {
  map_dfr(unique(weights_df$marker_group), function(grp) {
    w <- weights_df |> filter(marker_group == grp)
    genes_here <- intersect(w$gene, rownames(zmat))
    if (length(genes_here) == 0) return(NULL)
    w_vec <- w$weight[match(genes_here, w$gene)]
    score <- colSums(zmat[genes_here, , drop = FALSE] * w_vec) / sum(w_vec)
    tibble(sample_id = colnames(zmat), marker_group = grp, signature_score = score)
  })
}

incoming_signature_scores <- compute_signature_scores(incoming_full_z, signature_weights) |>
  left_join(
    incoming_sample_metadata |> select(sample_id, specimen_id, sample_class, maturation_week),
    by = "sample_id"
  ) |>
  filter(!is.na(maturation_week)) |>
  group_by(specimen_id, sample_class, maturation_week, marker_group) |>
  summarise(signature_score = mean(signature_score, na.rm = TRUE), .groups = "drop")

reference_signature_scores <- compute_signature_scores(reference_full_z, signature_weights) |>
  left_join(
    combined_metadata |> select(pseudobulk_id, reference_label, reference_level) |>
      rename(sample_id = pseudobulk_id),
    by = "sample_id"
  )

reference_signature_band <- reference_signature_scores |>
  filter(reference_level == "Neuronal subtype") |>
  group_by(marker_group) |>
  summarise(
    reference_mean = mean(signature_score, na.rm = TRUE),
    reference_low = quantile(signature_score, 0.1, na.rm = TRUE),
    reference_high = quantile(signature_score, 0.9, na.rm = TRUE),
    .groups = "drop"
  )

write.csv(
  incoming_signature_scores,
  "results/incoming_qc/weekly_gene_signature_scores.csv",
  row.names = FALSE
)

write.csv(
  reference_signature_band,
  "results/incoming_qc/pseudobulk_reference_signature_bands.csv",
  row.names = FALSE
)
```

## Weekly progression signal — condensed composite signature trajectory

One line per specimen per program, using the weighted composite signature
score above (not raw per-gene z-scores), plotted against the pseudobulk-derived
neuronal-subtype reference band. This is already better than the earlier
uniform-weight/shaded-band plot, but it should still be read as a
**program-by-program diagnostic view**, not the final maturity call, because
it spreads evidence across many facets and can visually overemphasize small
fluctuations in weak programs. The more reviewer-rigorous question is whether
a specimen shows a coherent *aggregate neuronal-identity gain* across weeks,
which is summarized in the section immediately below.

```{r}
#| label: fig-composite-signature-trajectory
#| fig-width: 13
#| fig-height: 8

ggplot(
  incoming_signature_scores,
  aes(x = maturation_week, y = signature_score, color = specimen_id)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_rect(
    data = reference_signature_band,
    aes(ymin = reference_low, ymax = reference_high),
    xmin = -Inf, xmax = Inf, fill = "grey40", alpha = 0.12, inherit.aes = FALSE
  ) +
  geom_hline(
    data = reference_signature_band,
    aes(yintercept = reference_mean),
    linetype = "dotted", color = "grey20", inherit.aes = FALSE
  ) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.4) +
  facet_wrap(~ marker_group, ncol = 3, scales = "free_y") +
  labs(
    title = "Replicate-condensed, correlation-weighted signature score by week",
    subtitle = "Grey band/dotted line = pseudobulk neuronal-subtype reference (10th-90th pct / mean)",
    x = "Maturation week",
    y = "Weighted composite signature score",
    color = "Specimen"
  ) +
  theme_classic()
```

## Aggregate neuronal identity index

To make weekly progression easier to review, collapse the five neuronal marker
programs into a single aggregate neuronal identity index and the five broad
non-neuronal programs into a single counter-index. This provides one compact,
reviewer-friendly maturity trajectory per specimen instead of asking the reader
to mentally integrate nine separate facets.

```{r}
#| label: fig-aggregate-neuronal-index
#| fig-width: 10
#| fig-height: 5

aggregate_identity_scores <- incoming_signature_scores |>
  mutate(
    program_class = case_when(
      marker_group %in% neuronal_programs ~ "neuronal",
      marker_group %in% non_neuronal_programs ~ "non_neuronal",
      TRUE ~ "other"
    )
  ) |>
  filter(program_class %in% c("neuronal", "non_neuronal")) |>
  group_by(specimen_id, sample_class, maturation_week, program_class) |>
  summarise(mean_signature_score = mean(signature_score, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = program_class, values_from = mean_signature_score) |>
  mutate(neuronal_identity_index = neuronal - non_neuronal)

ggplot(
  aggregate_identity_scores,
  aes(x = maturation_week, y = neuronal_identity_index, color = specimen_id)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_line(linewidth = 1) +
  geom_point(size = 2.6) +
  labs(
    title = "Aggregate neuronal identity index by week",
    subtitle = "Computed as mean neuronal-program signature minus mean non-neuronal-program signature",
    x = "Maturation week",
    y = "Neuronal identity index",
    color = "Specimen"
  ) +
  theme_classic()

write.csv(
  aggregate_identity_scores,
  "results/incoming_qc/aggregate_neuronal_identity_index.csv",
  row.names = FALSE
)
```

A reviewer should be able to answer the main biological question from this
single panel: does each specimen move toward stronger neuronal identity over
weeks, or does it plateau after an early gain? Use the facet-level program plot
above only to explain *why* a specimen's aggregate index rises or stalls.


## PCA-based neuronal identity trajectory

PCA is computed on the pseudobulk reference alone (defining "neuronal
identity space" from single-cell-derived ground truth), then incoming samples
are projected into that fixed space using the reference loadings. This is the
right statistical framing, but the default visualization can still be hard to
read because the incoming weekly points sit in a dense reference cloud and the
legend becomes overloaded. The plot below is therefore tuned explicitly for
reviewability: reference points are de-emphasized, incoming trajectories are
thickened and labelled, and specimen-week centroids are shown more prominently
than the underlying reference scatter.

```{r}
#| label: fig-pca-projection-trajectory
#| fig-width: 12
#| fig-height: 9

pca_genes <- projection_genes

reference_pca_input <- t(joint_logcpm[pca_genes, reference_ids, drop = FALSE])
reference_center <- colMeans(reference_pca_input)
reference_pca <- prcomp(reference_pca_input, center = TRUE, scale. = FALSE)

incoming_pca_input <- t(joint_logcpm[pca_genes, incoming_ids, drop = FALSE])
incoming_pca_input_centered <- sweep(incoming_pca_input, 2, reference_center, "-")
incoming_pca_projected <- incoming_pca_input_centered %*% reference_pca$rotation[, 1:2]

reference_pca_var <- 100 * summary(reference_pca)$importance[2, 1:2]

reference_pca_df <- combined_metadata |>
  transmute(
    sample_id = pseudobulk_id,
    reference_label = as.character(reference_label),
    reference_level = as.character(reference_level),
    PC1 = reference_pca$x[pseudobulk_id, 1],
    PC2 = reference_pca$x[pseudobulk_id, 2]
  )

incoming_pca_df <- incoming_sample_metadata |>
  filter(!is.na(maturation_week)) |>
  transmute(
    sample_id,
    specimen_id,
    sample_class,
    maturation_week,
    PC1 = incoming_pca_projected[sample_id, 1],
    PC2 = incoming_pca_projected[sample_id, 2]
  ) |>
  group_by(specimen_id, sample_class, maturation_week) |>
  summarise(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop") |>
  arrange(specimen_id, maturation_week)

reference_centroids <- reference_pca_df |>
  group_by(reference_label, reference_level) |>
  summarise(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot() +
  geom_point(
    data = reference_pca_df,
    aes(x = PC1, y = PC2, shape = reference_level),
    color = "grey75", alpha = 0.45, size = 2.2
  ) +
  geom_text(
    data = reference_centroids,
    aes(x = PC1, y = PC2, label = reference_label),
    color = "grey35", size = 3.2
  ) +
  geom_path(
    data = incoming_pca_df,
    aes(x = PC1, y = PC2, group = specimen_id, color = specimen_id),
    linewidth = 1.0,
    arrow = arrow(length = unit(0.18, "cm"), type = "closed")
  ) +
  geom_point(
    data = incoming_pca_df,
    aes(x = PC1, y = PC2, color = specimen_id, fill = maturation_week),
    shape = 21, size = 3.8, stroke = 0.5
  ) +
  ggrepel::geom_text_repel(
    data = incoming_pca_df |>
      group_by(specimen_id) |>
      slice_max(order_by = maturation_week, n = 1, with_ties = FALSE) |>
      ungroup(),
    aes(x = PC1, y = PC2, label = specimen_id, color = specimen_id),
    size = 3.2,
    show.legend = FALSE,
    max.overlaps = Inf
  ) +
  scale_fill_viridis_c() +
  labs(
    title = "Weekly maturation trajectory projected onto the pseudobulk reference PCA space",
    subtitle = "Grey points = reference pseudobulks; labels = reference centroids; colored arrows = specimen-week trajectory",
    x = sprintf("PC1 (%.1f%% ref variance)", reference_pca_var[1]),
    y = sprintf("PC2 (%.1f%% ref variance)", reference_pca_var[2]),
    color = "Specimen",
    shape = "Reference level",
    fill = "Week"
  ) +
  theme_classic()

write.csv(
  incoming_pca_df,
  "results/incoming_qc/weekly_pca_trajectory_coordinates.csv",
  row.names = FALSE
)
```

An arrow path that moves steadily toward the neuronal-subtype cluster over
weeks is the clearest single confirmation of on-target maturation this
notebook can produce; a path that stalls, reverses, or moves toward a
non-neuronal cluster is the strongest available signal for a protocol flag.

## Deconvolution — estimating cell-identity composition per sample

The single-cell-derived pseudobulk reference doubles as a signature matrix.
Non-negative least squares (NNLS) deconvolution regresses each incoming
sample's expression on the pseudobulk reference profiles (restricted to
reference-specific marker genes and the shared gene set), producing a
non-negative composition estimate per identity that sums to 1. This is
explicitly framed as a transcriptional-purity/composition estimate, not a
literal single-cell-resolution cell count, matching the caveat already used
for the broad TG pseudobulk reference in `TG_sc_reprocessing-3.qmd`.

Reviewer caution: in the current rendering, the NNLS output appears highly
degenerate, with most TG-bulk specimen-weeks collapsing almost entirely onto
neuronal identities and with NF2 dominating the subtype fraction. That may be
biologically partly true, but it is also a classic sign that the signature
matrix is too small, too collinear, or too marker-restricted for stable
subtype-level decomposition. Treat subtype fractions as *suggestive*, but rely
more heavily on the aggregate neuronal-vs-non-neuronal split unless the
signature matrix is expanded and benchmarked.

```{r}
#| label: nnls-deconvolution

if (!requireNamespace("nnls", quietly = TRUE)) {
  install.packages("nnls", repos = "https://cloud.r-project.org")
}
library(nnls)

deconv_signature_genes <- combined_marker_manifest |>
  filter(gene %in% rownames(joint_logcpm)) |>
  pull(gene) |>
  unique()

signature_matrix <- 2^joint_logcpm[deconv_signature_genes, reference_ids, drop = FALSE]

signature_identity_map <- combined_metadata |>
  transmute(pseudobulk_id, reference_label = as.character(reference_label)) |>
  filter(pseudobulk_id %in% colnames(signature_matrix))

identity_levels <- unique(signature_identity_map$reference_label)

signature_matrix_by_identity <- sapply(identity_levels, function(id_lab) {
  cols <- signature_identity_map$pseudobulk_id[signature_identity_map$reference_label == id_lab]
  if (length(cols) == 1) return(signature_matrix[, cols])
  rowMeans(signature_matrix[, cols, drop = FALSE])
})
colnames(signature_matrix_by_identity) <- identity_levels

incoming_linear_expr <- 2^joint_logcpm[deconv_signature_genes, incoming_ids, drop = FALSE]

run_nnls_fraction <- function(sample_expr, sig_matrix) {
  fit <- nnls::nnls(as.matrix(sig_matrix), as.numeric(sample_expr))
  coefs <- pmax(fit$x, 0)
  if (sum(coefs) == 0) coefs <- rep(0, length(coefs))
  else coefs <- coefs / sum(coefs)
  names(coefs) <- colnames(sig_matrix)
  coefs
}

deconv_fractions <- apply(incoming_linear_expr, 2, run_nnls_fraction, sig_matrix = signature_matrix_by_identity)
deconv_fractions <- as.data.frame(t(deconv_fractions)) |>
  rownames_to_column("sample_id")

deconv_long <- deconv_fractions |>
  pivot_longer(-sample_id, names_to = "reference_identity", values_to = "estimated_fraction") |>
  left_join(
    incoming_sample_metadata |> select(sample_id, specimen_id, sample_class, maturation_week),
    by = "sample_id"
  ) |>
  filter(!is.na(maturation_week)) |>
  group_by(specimen_id, sample_class, maturation_week, reference_identity) |>
  summarise(estimated_fraction = mean(estimated_fraction, na.rm = TRUE), .groups = "drop")

write.csv(
  deconv_long,
  "results/incoming_qc/weekly_nnls_deconvolution_fractions.csv",
  row.names = FALSE
)

knitr::kable(
  deconv_long |> pivot_wider(names_from = reference_identity, values_from = estimated_fraction),
  digits = 2,
  caption = "NNLS-estimated composition (fraction) per specimen x week, condensed across replicates"
)
```

```{r}
#| label: fig-deconvolution-stacked-bar
#| fig-width: 13
#| fig-height: 7

ggplot(
  deconv_long,
  aes(x = factor(maturation_week), y = estimated_fraction, fill = reference_identity)
) +
  geom_col(position = "stack") +
  facet_wrap(~ specimen_id, ncol = 3) +
  labs(
    title = "NNLS-estimated transcriptional composition by week",
    subtitle = "Signature matrix = pseudobulk reference identities restricted to curated marker genes",
    x = "Maturation week",
    y = "Estimated fraction",
    fill = "Reference identity"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Deconvolution — broad identity summary

Because subtype-level NNLS is vulnerable to collinearity among closely related
neuronal programs (e.g. NF2 vs NF3 vs PEP), also collapse the deconvolution to
a broad reviewer-facing readout: neuronal vs non-neuronal composition by week.
This is the more stable interpretation layer for acceptance/QC decisions.

```{r}
#| label: fig-deconvolution-broad-summary
#| fig-width: 10
#| fig-height: 5

broad_deconv_summary <- deconv_long |>
  mutate(
    broad_identity = ifelse(reference_identity %in% neuronal_identity_labels, "Neuronal", "Non-neuronal")
  ) |>
  group_by(specimen_id, sample_class, maturation_week, broad_identity) |>
  summarise(estimated_fraction = sum(estimated_fraction), .groups = "drop")

ggplot(
  broad_deconv_summary,
  aes(x = maturation_week, y = estimated_fraction, color = specimen_id)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ broad_identity, ncol = 1) +
  labs(
    title = "Broad deconvolution summary by week",
    subtitle = "More stable than subtype-level NNLS for reviewer-facing interpretation",
    x = "Maturation week",
    y = "Estimated fraction",
    color = "Specimen"
  ) +
  theme_classic()

write.csv(
  broad_deconv_summary,
  "results/incoming_qc/weekly_nnls_broad_fractions.csv",
  row.names = FALSE
)
```


Rising neuronal-subtype fraction (and falling non-neuronal fraction) across
weeks within a specimen is the deconvolution-based counterpart to the PCA
trajectory and composite signature score above; agreement across all three
methods (signature score, PCA projection, and deconvolution fraction) is the
strongest available evidence for genuine maturation, while disagreement
between them flags a case worth manual review rather than an automatic
accept/reject.

## Week-over-week maturation flags (condensed signal)

Flags are now computed on the **replicate-condensed, correlation-weighted
composite signature score** and the **NNLS neuronal-subtype fraction**,
rather than an unweighted per-gene z-score, and require agreement between the
two before raising a flag — reducing false positives driven by a single noisy
program or gene.

```{r}
#| label: week-over-week-flags

neuronal_programs <- c(
  "Pan-neuronal", "Peptidergic nociceptor", "Non-peptidergic nociceptor",
  "Neurofilament / mechanosensory", "Cold / pruriceptor-associated"
)

non_neuronal_programs <- c("Satellite glia", "Schwann cell", "Fibroblast", "Vascular", "Immune")

signature_flags <- incoming_signature_scores |>
  arrange(specimen_id, marker_group, maturation_week) |>
  group_by(specimen_id, marker_group) |>
  mutate(
    delta_vs_prior_week = signature_score - lag(signature_score),
    program_class = case_when(
      marker_group %in% neuronal_programs ~ "neuronal",
      marker_group %in% non_neuronal_programs ~ "non_neuronal",
      TRUE ~ "other"
    )
  ) |>
  ungroup() |>
  filter(!is.na(delta_vs_prior_week)) |>
  mutate(
    stalled_neuronal_flag = program_class == "neuronal" & delta_vs_prior_week <= 0,
    rising_non_neuronal_flag = program_class == "non_neuronal" & delta_vs_prior_week > 0.15
  )

signature_flag_summary <- signature_flags |>
  group_by(specimen_id, maturation_week) |>
  summarise(
    any_neuronal_program_stalled = any(stalled_neuronal_flag),
    any_non_neuronal_program_rising = any(rising_non_neuronal_flag),
    .groups = "drop"
  )

neuronal_identity_labels <- combined_metadata |>
  filter(reference_level == "Neuronal subtype") |>
  pull(reference_label) |>
  as.character() |>
  unique()

deconv_neuronal_fraction <- deconv_long |>
  mutate(is_neuronal_identity = reference_identity %in% neuronal_identity_labels) |>
  group_by(specimen_id, sample_class, maturation_week) |>
  summarise(neuronal_fraction = sum(estimated_fraction[is_neuronal_identity]), .groups = "drop") |>
  arrange(specimen_id, maturation_week) |>
  group_by(specimen_id) |>
  mutate(delta_neuronal_fraction = neuronal_fraction - lag(neuronal_fraction)) |>
  ungroup()

combined_flag_check <- signature_flags |>
  filter(program_class == "neuronal") |>
  group_by(specimen_id, maturation_week) |>
  summarise(any_neuronal_program_stalled = any(stalled_neuronal_flag), .groups = "drop") |>
  left_join(deconv_neuronal_fraction, by = c("specimen_id", "maturation_week")) |>
  mutate(
    deconv_stalled = !is.na(delta_neuronal_fraction) & delta_neuronal_fraction <= 0,
    concordant_stall_flag = any_neuronal_program_stalled & deconv_stalled
  )

knitr::kable(
  combined_flag_check |> filter(concordant_stall_flag),
  digits = 3,
  caption = "Specimen-weeks where BOTH the signature score and NNLS neuronal fraction agree maturation stalled"
)

write.csv(
  combined_flag_check,
  "results/incoming_qc/weekly_maturation_flags_condensed.csv",
  row.names = FALSE
)
```

## Weekly intake summary

Before interpreting the summary counts below, note that the biologically
meaningful unit is the **specimen-week**, not the individual marker program.
The counts here are therefore specimen-week summaries, avoiding inflation from
multiple correlated neuronal programs all flagging the same sample-week.

```{r}
#| label: weekly-intake-summary

intake_summary <- tibble(
  metric = c(
    "Incoming samples this run",
    "Samples flagged on depth/complexity QC",
    "Samples with identity mismatch vs. expected neuronal reference",
    "Specimen/week combinations with any stalled neuronal signature program",
    "Specimen/week combinations with a rising non-neuronal signature program",
    "Specimen/week combinations with concordant signature + NNLS stall flag"
  ),
  value = c(
    nrow(incoming_sample_metadata),
    sum(incoming_sample_metadata$qc_flag),
    sum(incoming_expect_neuron_flag$identity_flag),
    sum(signature_flag_summary$any_neuronal_program_stalled, na.rm = TRUE),
    sum(signature_flag_summary$any_non_neuronal_program_rising, na.rm = TRUE),
    sum(combined_flag_check$concordant_stall_flag, na.rm = TRUE)
  )
)

knitr::kable(intake_summary)
```

## Next steps checklist

- Any sample flagged in **Incoming sample QC** should be re-checked against
  the wet-lab log (RIN, input mass, library prep batch) before being pooled
  into any comparison.
- Keep the **identity-mismatch table** focused on true off-target controls
  (`hDRG`, `TAK2204464A`) versus intended TG-bulk maturation samples; those
  are useful positive controls for the reference anchoring logic and should
  not be conflated with failures of the TG differentiation series.
- Use the **aggregate neuronal identity index** as the primary reviewer-facing
  maturity readout, then use the per-program signature facets to explain which
  component programs drive a rise, plateau, or reversal.
- Use the **PCA trajectory plot** as the main geometric readout of neuronal
  identity progression: the question is whether specimen centroids move toward
  the neuronal reference manifold over weeks, not whether any single week is
  closest to one subtype label.
- Treat **subtype-level NNLS fractions** as exploratory because the current
  signature matrix appears collinear; prefer the broad neuronal vs
  non-neuronal deconvolution summary for acceptance/QC decisions until the
  subtype matrix is benchmarked.
- Any specimen/week combination flagged as **concordant** (both the
  composite signature score and the broad neuronal fraction stall or reverse)
  should trigger a note in the differentiation log; two consecutive
  concordant-stall weeks for the same specimen is the current trigger for
  re-derivation review.
- Revisit the correlation-based gene weights and the NNLS signature gene set
  periodically as more donors/timepoints accumulate in the pseudobulk
  reference, since both are learned from that reference and will sharpen as
  it grows.
