---
title: "Incoming RNA-seq Sample QC & Weekly Maturation Tracking"
subtitle: "TG-reference-anchored review of new bulk RNA-seq batches (RealTGN.salmon)"
author: "Austin Hovland"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Purpose

This notebook is the recurring intake-QC and maturation-tracking companion to `TG_sc_reprocessing-3.qmd` (single-nucleus reference build) and `TG_reviewer_figures.qmd` (bulk-to-reference projection). It is meant to be re-run every time a new batch of bulk RNA-seq quantifications lands, and does two things:

1.  **Incoming sample QC** — library size, complexity, duplication/composition sanity checks, and similarity to the integrated TG single-nucleus reference, flagging any sample that looks degraded, contaminated, or mislabeled before it enters downstream analysis.
2.  **Weekly maturation investigation** — tracks the curated marker sets already derived for the TG reference (pan-neuronal, peptidergic/nociceptor, non-peptidergic, neurofilament/mechanosensory, cold/pruriceptor, and the non-neuronal purity programs) across the differentiation time course encoded in each sample's well/week label (e.g. `TGN...W1`, `W2`, ...), so drift or stalled maturation is caught early. Every marker-based comparison is explicitly **anchored to the single-cell-derived pseudobulk reference**: incoming samples are jointly normalized with the pseudobulk profiles, plotted on the same heatmap and z-score scale as those profiles, and evaluated against a pseudobulk-derived reference band rather than against the incoming cohort alone.

This notebook assumes `TG_sc_reprocessing-3.qmd` has already been run (so `results/pseudobulk/` is populated) and that new bulk counts are available as a Salmon merged gene-counts TSV, following the same layout as `RealTGN.salmon.merged.gene_counts.tsv`.

## Setup

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(edgeR)

dir.create("results/incoming_qc", recursive = TRUE, showWarnings = FALSE)

# --- Point this at the newest merged Salmon gene-counts table each week ---
incoming_counts_path <- "RealTGN.salmon.merged.gene_counts.tsv"

# Reference artifacts produced by TG_sc_reprocessing-3.qmd
combined_dge <- readRDS(
  "results/pseudobulk/GSE197289_TG_combined_reference_edgeR_dge.rds"
)

combined_logcpm <- as.matrix(
  read.csv(
    "results/pseudobulk/GSE197289_TG_combined_reference_logCPM.csv",
    row.names = 1,
    check.names = FALSE
  )
)

combined_metadata <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_reference_metadata.csv",
  show_col_types = FALSE
)

# Previously derived marker manifest (pan-neuronal, subtype, non-neuronal programs)
combined_marker_manifest <- read_csv(
  "results/pseudobulk/GSE197289_TG_combined_marker_manifest.csv",
  show_col_types = FALSE
)

stopifnot(identical(colnames(combined_logcpm), combined_metadata$pseudobulk_id))
```

## Import incoming bulk RNA-seq samples

```{r}
#| label: import-incoming-samples

incoming_quant <- read_tsv(
  incoming_counts_path,
  show_col_types = FALSE
)

stopifnot(all(c("gene_id", "gene_name") %in% colnames(incoming_quant)))

incoming_sample_ids <- setdiff(colnames(incoming_quant), c("gene_id", "gene_name"))

stopifnot(length(incoming_sample_ids) > 0L)

incoming_counts_tbl <- incoming_quant |>
  transmute(
    gene_name = toupper(trimws(gene_name)),
    across(all_of(incoming_sample_ids), as.numeric)
  ) |>
  filter(!is.na(gene_name), gene_name != "") |>
  group_by(gene_name) |>
  summarise(
    across(all_of(incoming_sample_ids), \(x) sum(x, na.rm = TRUE)),
    .groups = "drop"
  )

incoming_counts <- incoming_counts_tbl |> as.data.frame()
rownames(incoming_counts) <- incoming_counts$gene_name
incoming_counts$gene_name <- NULL
incoming_counts <- as.matrix(incoming_counts)

stopifnot(!anyDuplicated(rownames(incoming_counts)))
```

## Parse sample identity and week/maturation timepoint

Sample labels follow the established convention: cohort prefix, then `hDRG_<n>`, `TGN<label>W<week>`, or `TAK<lot>`. The trailing `W<number>` token is parsed out explicitly as the differentiation week for the maturation time course; samples without a parseable week are retained but excluded from the weekly trajectory plots.

```{r}
#| label: parse-sample-identity

incoming_total_counts <- colSums(incoming_counts)
incoming_detected_genes <- colSums(incoming_counts > 0)

incoming_sample_metadata <- tibble(
  sample_id = colnames(incoming_counts)
) |>
  mutate(
    cohort = sub("^([^_]+)_.*$", "\\1", sample_id),
    sample_label = sub("^[^_]+_", "", sample_id),
    sample_class = case_when(
      grepl("^hDRG", sample_label) ~ "hDRG",
      grepl("^TGN", sample_label) ~ "TG bulk",
      grepl("^TAK", sample_label) ~ "TAK2204464A",
      TRUE ~ "Other"
    ),
    specimen_id = case_when(
      grepl("^hDRG_[0-9]+$", sample_label) ~ sub("_[0-9]+$", "", sample_label),
      grepl("^TGN.*W[0-9]+$", sample_label) ~ sub("W[0-9]+$", "", sample_label),
      TRUE ~ sample_label
    ),
    week_raw = str_extract(sample_label, "W[0-9]+$"),
    maturation_week = as.integer(str_remove(week_raw, "^W")),
    total_counts = incoming_total_counts[sample_id],
    detected_genes = incoming_detected_genes[sample_id],
    genes_per_million_counts = 1e6 * detected_genes / total_counts
  )

n_missing_week <- sum(is.na(incoming_sample_metadata$maturation_week))

message(
  "Samples parsed: ", nrow(incoming_sample_metadata),
  "\nSamples without a parseable maturation week: ", n_missing_week
)

knitr::kable(
  incoming_sample_metadata |>
    select(sample_id, cohort, sample_class, specimen_id, maturation_week,
           total_counts, detected_genes),
  digits = 0
)

write.csv(
  incoming_sample_metadata,
  "results/incoming_qc/incoming_sample_metadata.csv",
  row.names = FALSE
)
```

## Incoming sample QC — library size and complexity

**QC thresholds** below mirror the conservative single-nucleus flags already established in `TG_sc_reprocessing-3.qmd` (`qc_min_counts`, `qc_min_features`) but rescaled for bulk libraries; adjust `bulk_min_total_counts` and `bulk_min_detected_genes` as the pipeline's bulk QC baseline is refined.

```{r}
#| label: incoming-qc-thresholds
#| fig-width: 11
#| fig-height: 5

bulk_min_total_counts <- 5e6
bulk_min_detected_genes <- 10000L

incoming_sample_metadata <- incoming_sample_metadata |>
  mutate(
    qc_low_depth = total_counts < bulk_min_total_counts,
    qc_low_complexity = detected_genes < bulk_min_detected_genes,
    qc_flag = qc_low_depth | qc_low_complexity,
    qc_status = factor(
      ifelse(qc_flag, "flagged", "retain"),
      levels = c("retain", "flagged")
    )
  )

knitr::kable(
  incoming_sample_metadata |>
    filter(qc_flag) |>
    select(sample_id, sample_class, total_counts, detected_genes,
           qc_low_depth, qc_low_complexity),
  digits = 0
)

ggplot(
  incoming_sample_metadata,
  aes(x = total_counts, y = detected_genes, color = sample_class, shape = qc_status)
) +
  geom_vline(xintercept = bulk_min_total_counts, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = bulk_min_detected_genes, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = sample_id), size = 2.6, max.overlaps = 20, show.legend = FALSE) +
  scale_x_log10() +
  labs(
    title = "Incoming bulk RNA-seq library depth vs. complexity",
    subtitle = "Dashed lines mark the current bulk QC floor",
    x = "Total counts (log scale)",
    y = "Detected genes",
    color = "Sample class",
    shape = "QC status"
  ) +
  theme_classic()

write.csv(
  incoming_sample_metadata,
  "results/incoming_qc/incoming_sample_qc_flags.csv",
  row.names = FALSE
)
```

## Joint normalization with the TG reference

Incoming samples are jointly TMM-normalized with the filtered combined TG pseudobulk reference over their shared gene set, exactly as in `TG_reviewer_figures.qmd`, so every weekly batch lands on the same expression scale as the reference before any similarity or marker comparison is made.

```{r}
#| label: joint-normalization

reference_gene_key <- toupper(rownames(combined_dge$counts))
stopifnot(!anyDuplicated(reference_gene_key))
stopifnot(!anyDuplicated(rownames(incoming_counts)))

shared_gene_key <- intersect(reference_gene_key, rownames(incoming_counts))

reference_counts_shared <- combined_dge$counts[
  match(shared_gene_key, reference_gene_key), , drop = FALSE
]
incoming_counts_shared <- incoming_counts[
  match(shared_gene_key, rownames(incoming_counts)), , drop = FALSE
]

rownames(reference_counts_shared) <- shared_gene_key
rownames(incoming_counts_shared) <- shared_gene_key

stopifnot(identical(rownames(reference_counts_shared), rownames(incoming_counts_shared)))
stopifnot(nrow(reference_counts_shared) > 10000L)

joint_counts <- cbind(reference_counts_shared, incoming_counts_shared)

joint_sample_metadata <- bind_rows(
  combined_metadata |>
    transmute(
      sample_id = pseudobulk_id,
      sample_origin = "Reference pseudobulk",
      reference_label = as.character(reference_label),
      reference_level = as.character(reference_level),
      cohort = paste0("donor_", donor),
      maturation_week = NA_integer_
    ),
  incoming_sample_metadata |>
    transmute(
      sample_id,
      sample_origin = "Incoming bulk RNA-seq",
      reference_label = sample_class,
      reference_level = specimen_id,
      cohort,
      maturation_week
    )
)

joint_sample_metadata <- joint_sample_metadata[
  match(colnames(joint_counts), joint_sample_metadata$sample_id), , drop = FALSE
]
stopifnot(identical(colnames(joint_counts), joint_sample_metadata$sample_id))

joint_dge <- DGEList(counts = joint_counts, samples = as.data.frame(joint_sample_metadata))
joint_dge <- calcNormFactors(joint_dge, method = "TMM")
joint_logcpm <- cpm(joint_dge, log = TRUE, prior.count = 2)

n_top_var_genes <- 2000
gene_var <- apply(combined_logcpm, 1, var)
top_var_genes <- names(sort(gene_var, decreasing = TRUE))[seq_len(min(n_top_var_genes, length(gene_var)))]
projection_genes <- intersect(toupper(top_var_genes), rownames(joint_logcpm))
stopifnot(length(projection_genes) >= 500L)
```

## Incoming sample similarity to TG reference identities

Flags samples whose strongest correlation is not to the expected neuronal program (e.g. contamination, mislabeling, or a maturation arrest that looks non-neuronal transcriptionally).

```{r}
#| label: incoming-reference-correlation
#| fig-width: 12
#| fig-height: 10

incoming_ids <- incoming_sample_metadata$sample_id
reference_ids <- combined_metadata$pseudobulk_id

incoming_reference_cor <- cor(
  joint_logcpm[projection_genes, incoming_ids, drop = FALSE],
  joint_logcpm[projection_genes, reference_ids, drop = FALSE],
  method = "spearman"
)

incoming_reference_long <- as.data.frame(as.table(incoming_reference_cor)) |>
  rename(bulk_sample = Var1, reference_id = Var2, spearman_correlation = Freq) |>
  left_join(
    combined_metadata |>
      transmute(
        reference_id = pseudobulk_id,
        reference_label = as.character(reference_label),
        reference_level = as.character(reference_level)
      ),
    by = "reference_id"
  )

incoming_reference_summary <- incoming_reference_long |>
  group_by(bulk_sample, reference_label, reference_level) |>
  summarise(median_correlation = median(spearman_correlation), .groups = "drop") |>
  arrange(bulk_sample, desc(median_correlation))

incoming_top_matches <- incoming_reference_summary |>
  group_by(bulk_sample) |>
  slice_max(order_by = median_correlation, n = 3, with_ties = FALSE) |>
  ungroup()

knitr::kable(incoming_top_matches, digits = 3)

incoming_expect_neuron_flag <- incoming_top_matches |>
  group_by(bulk_sample) |>
  slice_max(order_by = median_correlation, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    sample_id = as.character(bulk_sample),
    top_reference_match = reference_label,
    top_reference_level = reference_level,
    top_match_correlation = median_correlation,
    identity_flag = !top_reference_level %in% c("Neuronal subtype")
  )

knitr::kable(
  incoming_expect_neuron_flag |> filter(identity_flag),
  digits = 3
)

write.csv(
  incoming_reference_summary,
  "results/incoming_qc/incoming_bulk_reference_correlation.csv",
  row.names = FALSE
)

write.csv(
  incoming_expect_neuron_flag,
  "results/incoming_qc/incoming_identity_flags.csv",
  row.names = FALSE
)
```

## Weekly maturation investigation — curated marker trajectories

This reuses the marker sets already curated and validated against the TG single-nucleus reference (`combined_marker_manifest`, from `TG_sc_reprocessing-3.qmd`) rather than redefining a new panel. Each gene's log-CPM in the jointly normalized space is plotted against `maturation_week` per specimen, faceted by marker program, so week-over-week gain of neuronal/subtype identity (and absence of non-neuronal drift) is visible at a glance.

```{r}
#| label: maturation-marker-trajectories
#| fig-width: 14
#| fig-height: 12

maturation_markers_present <- combined_marker_manifest |>
  filter(gene %in% rownames(joint_logcpm))

maturation_expr_long <- joint_logcpm[maturation_markers_present$gene, incoming_ids, drop = FALSE] |>
  as.data.frame() |>
  rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "sample_id", values_to = "log_cpm") |>
  left_join(maturation_markers_present, by = "gene") |>
  left_join(
    incoming_sample_metadata |> select(sample_id, specimen_id, sample_class, maturation_week),
    by = "sample_id"
  ) |>
  filter(!is.na(maturation_week))

ggplot(
  maturation_expr_long,
  aes(x = maturation_week, y = log_cpm, color = marker_group, group = paste(gene, specimen_id))
) +
  geom_line(alpha = 0.35) +
  geom_point(size = 1.6, alpha = 0.6) +
  facet_wrap(~ marker_group, scales = "free_y", ncol = 3) +
  labs(
    title = "Marker-set expression across differentiation week",
    subtitle = "Curated TG reference marker programs, jointly normalized with the single-nucleus reference",
    x = "Maturation week",
    y = "log2 CPM (joint TMM)",
    color = "Marker program"
  ) +
  theme_classic() +
  theme(legend.position = "none")

write.csv(
  maturation_expr_long,
  "results/incoming_qc/weekly_maturation_marker_expression.csv",
  row.names = FALSE
)
```

## Weekly maturation summary score per program

Collapses each marker program to a per-sample mean z-scored log-CPM, giving one maturation trajectory line per program per specimen. To anchor these trajectories against the single-cell-derived pseudobulk reference rather than judging them in isolation, the z-scores are computed in the same jointly normalized space as the pseudobulk reference, and the trajectory plot below overlays each program's expected pseudobulk neuronal-subtype reference band (10th-90th percentile) so week-over-week progress can be read directly against where mature single-cell-derived neuronal identity actually sits.

```{r}
#| label: maturation-program-score
#| fig-width: 12
#| fig-height: 7

program_matrix <- joint_logcpm[maturation_markers_present$gene, incoming_ids, drop = FALSE]
program_matrix_z <- t(scale(t(program_matrix)))
program_matrix_z <- program_matrix_z[apply(program_matrix_z, 1, function(x) all(is.finite(x))), , drop = FALSE]

program_score_long <- program_matrix_z |>
  as.data.frame() |>
  rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "sample_id", values_to = "z_score") |>
  left_join(maturation_markers_present, by = "gene") |>
  left_join(
    incoming_sample_metadata |> select(sample_id, specimen_id, sample_class, maturation_week),
    by = "sample_id"
  ) |>
  filter(!is.na(maturation_week)) |>
  group_by(specimen_id, sample_class, maturation_week, marker_group) |>
  summarise(mean_z_score = mean(z_score, na.rm = TRUE), .groups = "drop")

reference_program_matrix <- joint_logcpm[maturation_markers_present$gene, reference_ids, drop = FALSE]
reference_program_matrix_z <- t(scale(t(reference_program_matrix)))
reference_program_matrix_z <- reference_program_matrix_z[
  apply(reference_program_matrix_z, 1, function(x) all(is.finite(x))),
  ,
  drop = FALSE
]

reference_program_score_long <- reference_program_matrix_z |>
  as.data.frame() |>
  rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "pseudobulk_id", values_to = "z_score") |>
  left_join(maturation_markers_present, by = "gene") |>
  left_join(
    combined_metadata |> select(pseudobulk_id, reference_label, reference_level),
    by = "pseudobulk_id"
  ) |>
  mutate(
    is_neuronal_reference = reference_level == "Neuronal subtype"
  )

reference_program_band <- reference_program_score_long |>
  filter(is_neuronal_reference) |>
  group_by(marker_group) |>
  summarise(
    reference_mean_z = mean(z_score, na.rm = TRUE),
    reference_low = quantile(z_score, 0.1, na.rm = TRUE),
    reference_high = quantile(z_score, 0.9, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  program_score_long,
  aes(x = maturation_week, y = mean_z_score, color = marker_group)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_rect(
    data = reference_program_band,
    aes(ymin = reference_low, ymax = reference_high, fill = marker_group),
    xmin = -Inf, xmax = Inf, alpha = 0.08, inherit.aes = FALSE, show.legend = FALSE
  ) +
  geom_hline(
    data = reference_program_band,
    aes(yintercept = reference_mean_z, color = marker_group),
    linetype = "dotted", linewidth = 0.5, show.legend = FALSE
  ) +
  geom_line(aes(group = interaction(specimen_id, marker_group)), alpha = 0.6) +
  geom_point(size = 2.2) +
  facet_wrap(~ specimen_id, ncol = 3) +
  labs(
    title = "Per-program maturation score by week and specimen, anchored to the pseudobulk reference",
    subtitle = "Shaded band = 10th-90th percentile of the matching neuronal-subtype pseudobulk reference; dotted line = reference mean",
    x = "Maturation week",
    y = "Mean marker-program z-score",
    color = "Marker program"
  ) +
  theme_classic()

write.csv(
  reference_program_band,
  "results/incoming_qc/pseudobulk_reference_program_bands.csv",
  row.names = FALSE
)

program_score_wide <- program_score_long |>
  pivot_wider(names_from = marker_group, values_from = mean_z_score)

knitr::kable(program_score_wide, digits = 2)

knitr::kable(reference_program_band, digits = 2, caption = "Neuronal-subtype pseudobulk reference band per marker program (this is the single-cell-derived anchor incoming samples are plotted against above)")

write.csv(
  program_score_wide,
  "results/incoming_qc/weekly_maturation_program_scores.csv",
  row.names = FALSE
)
```

## Week-over-week maturation flags

Automatic screen for the two most common weekly QC failure modes: (1) a neuronal program score that fails to increase between consecutive weeks within the same specimen (stalled maturation), and (2) a non-neuronal program (fibroblast, glial, immune, vascular) rising alongside the neuronal score (possible contamination or off-target differentiation).

```{r}
#| label: week-over-week-flags

neuronal_programs <- c(
  "Pan-neuronal", "Peptidergic nociceptor", "Non-peptidergic nociceptor",
  "Neurofilament / mechanosensory", "Cold / pruriceptor-associated"
)

non_neuronal_programs <- c("Satellite glia", "Schwann cell", "Fibroblast", "Vascular", "Immune")

maturation_flags <- program_score_long |>
  arrange(specimen_id, marker_group, maturation_week) |>
  group_by(specimen_id, marker_group) |>
  mutate(
    delta_vs_prior_week = mean_z_score - lag(mean_z_score),
    program_class = case_when(
      marker_group %in% neuronal_programs ~ "neuronal",
      marker_group %in% non_neuronal_programs ~ "non_neuronal",
      TRUE ~ "other"
    )
  ) |>
  ungroup() |>
  filter(!is.na(delta_vs_prior_week)) |>
  mutate(
    stalled_neuronal_flag = program_class == "neuronal" & delta_vs_prior_week <= 0,
    rising_non_neuronal_flag = program_class == "non_neuronal" & delta_vs_prior_week > 0.25
  )

knitr::kable(
  maturation_flags |> filter(stalled_neuronal_flag | rising_non_neuronal_flag),
  digits = 2
)

write.csv(
  maturation_flags,
  "results/incoming_qc/weekly_maturation_flags.csv",
  row.names = FALSE
)
```

## Weekly intake summary

```{r}
#| label: weekly-intake-summary

intake_summary <- tibble(
  metric = c(
    "Incoming samples this run",
    "Samples flagged on depth/complexity QC",
    "Samples with identity mismatch vs. expected neuronal reference",
    "Specimen/week combinations with a stalled-neuronal flag",
    "Specimen/week combinations with a rising non-neuronal flag"
  ),
  value = c(
    nrow(incoming_sample_metadata),
    sum(incoming_sample_metadata$qc_flag),
    sum(incoming_expect_neuron_flag$identity_flag),
    sum(maturation_flags$stalled_neuronal_flag),
    sum(maturation_flags$rising_non_neuronal_flag)
  )
)

knitr::kable(intake_summary)
```

## Next steps checklist

- Any sample flagged in **Incoming sample QC** should be re-checked against the wet-lab log (RIN, input mass, library prep batch) before being pooled into any comparison.
- Any sample flagged in **Incoming sample similarity to TG reference identities** with `identity_flag = TRUE` should be visually re-inspected in the full similarity heatmap and the anchored marker heatmap in this notebook (and cross-checked against the PCA/heatmap from `TG_reviewer_figures.qmd`) before acceptance.
- Use the **pseudobulk reference bands** in the per-program maturation score plot as the primary pass/fail anchor for maturation progress — a specimen tracking toward (not just away from zero, but into) the neuronal-subtype reference band by week 4 is the target trajectory; one that plateaus well outside the band warrants a differentiation-protocol review.
- Any specimen/week combination in **Week-over-week maturation flags** should trigger a note in the differentiation log; two consecutive stalled weeks for the same specimen is the current trigger for re-derivation review.
