---
title: "Human Trigeminal Ganglion Reference: GSE197289"
author: "Austin Hovland"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Objective

Construct a reproducible Seurat v5 object from the processed human TG snRNA-seq
counts and barcode metadata released with GSE197289.

The source matrix is genes × nuclei. The metadata file contains the cell barcode
in `V1`, which is renamed to `barcode` below. The `cellID` field is an
author-provided label and is not unique.

## Setup

```{r}
#| label: setup

library(Seurat)
library(SeuratObject)
library(Matrix)
library(readr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(glmGamPoi)
library(edgeR)
library(pheatmap)
library(ggrepel)

set.seed(197289)

counts_path <- "GSE197289_snRNA-seq_human_raw_counts.RDS/GSE197289_snRNA-seq_human_raw_counts.RDS"
meta_path <- "GSE197289_snRNA-seq_human_barcode_meta.csv/GSE197289_snRNA-seq_human_barcode_meta.csv"

dir.create("results", showWarnings = FALSE)
dir.create("results/qc", recursive = TRUE, showWarnings = FALSE)
dir.create("results/seurat", recursive = TRUE, showWarnings = FALSE)
dir.create("results/pseudobulk", recursive = TRUE, showWarnings = FALSE)

package_versions <- tibble(
  package = c(
    "Seurat", "SeuratObject", "Matrix", "readr", "dplyr",
    "ggplot2", "patchwork", "glmGamPoi", "edgeR",
    "pheatmap", "ggrepel"
  ),
  version = vapply(
    c(
      "Seurat", "SeuratObject", "Matrix", "readr", "dplyr",
      "ggplot2", "patchwork", "glmGamPoi", "edgeR",
      "pheatmap", "ggrepel"
    ),
    \(pkg) as.character(packageVersion(pkg)),
    character(1)
  )
)

knitr::kable(package_versions)
```

## Import and validate

```{r}
#| label: import-validate

counts <- readRDS(counts_path)

meta <- read_csv(
  meta_path,
  show_col_types = FALSE
) |>
  rename(barcode = V1) |>
  as.data.frame()

stopifnot(inherits(counts, "dgCMatrix"))
stopifnot(!is.null(rownames(counts)))
stopifnot(!is.null(colnames(counts)))
stopifnot(!anyDuplicated(rownames(counts)))
stopifnot(!anyDuplicated(colnames(counts)))

stopifnot("barcode" %in% colnames(meta))
stopifnot(!anyDuplicated(meta$barcode))
stopifnot(ncol(counts) == nrow(meta))
stopifnot(setequal(colnames(counts), meta$barcode))

dataset_dimensions <- tibble(
  metric = c("Genes", "Nuclei", "Metadata rows", "Unique count-matrix barcodes"),
  value = c(
    nrow(counts),
    ncol(counts),
    nrow(meta),
    length(unique(colnames(counts)))
  )
)

knitr::kable(dataset_dimensions)
```

## Align metadata

```{r}
#| label: align-metadata

barcode_idx <- match(colnames(counts), meta$barcode)

stopifnot(!anyNA(barcode_idx))
stopifnot(length(barcode_idx) == ncol(counts))

# dplyr::slice avoids RStudio's occasional issue with `df[i, , drop = FALSE]`
meta <- dplyr::slice(meta, barcode_idx)
meta <- as.data.frame(meta)

stopifnot(identical(meta$barcode, colnames(counts)))

rownames(meta) <- meta$barcode

stopifnot(identical(rownames(meta), colnames(counts)))

knitr::kable(
  meta |>
    select(
      barcode, ganglia, class, cellID, subtype,
      donor, library, nCount_RNA, nFeature_RNA, percent.mt
    ) |>
    head(10)
)
```

## Create Seurat object

```{r}
#| label: create-seurat-object

tg <- CreateSeuratObject(
  counts = counts,
  meta.data = meta,
  project = "GSE197289_human_TG",
  assay = "RNA",
  min.cells = 0,
  min.features = 0
)

# Preserve original GEO annotation and supplied QC fields.
tg$author_label_short <- tg$cellID
tg$cell_type_author <- factor(tg$class)
tg$subtype_author <- factor(tg$subtype)
tg$percent_mt_author <- tg$percent.mt
tg$nCount_RNA_author <- tg$nCount_RNA
tg$nFeature_RNA_author <- tg$nFeature_RNA

# Declare covariates.
tg$donor <- factor(tg$donor)
tg$library <- factor(tg$library)
tg$ganglia <- factor(tg$ganglia)

# Required data-integrity checks.
stopifnot(identical(colnames(tg), rownames(tg[[]])))
stopifnot(ncol(tg) == 38028L)
stopifnot(nrow(tg) == 32214L)

print(tg)
```
## Annotation audit

```{r}
#| label: annotation-audit

broad_annotation <- tg[[]] |>
  count(cell_type_author, name = "n_nuclei") |>
  arrange(desc(n_nuclei))

subtype_annotation <- tg[[]] |>
  count(subtype_author, name = "n_nuclei") |>
  arrange(desc(n_nuclei))

donor_by_class <- tg[[]] |>
  count(donor, cell_type_author, name = "n_nuclei") |>
  arrange(donor, desc(n_nuclei))

knitr::kable(broad_annotation)
knitr::kable(subtype_annotation)
knitr::kable(donor_by_class)

write.csv(
  subtype_annotation,
  "results/qc/GSE197289_subtype_author_counts.csv",
  row.names = FALSE
)
```

## Recalculate QC metrics

The original metadata contains author-provided counts, detected features, and
mitochondrial percentage. Recalculate mitochondrial, ribosomal, and hemoglobin
percentages from the supplied matrix for audit purposes. No cells are removed
at this stage.

```{r}
#| label: qc-metrics

tg[["percent.mt"]] <- PercentageFeatureSet(
  tg,
  pattern = "^MT-"
)

tg[["percent.ribo"]] <- PercentageFeatureSet(
  tg,
  pattern = "^RP[SL]"
)

hb_genes <- intersect(
  c("HBA1", "HBA2", "HBB", "HBD", "HBE1", "HBG1", "HBG2", "HBM", "HBQ1", "HBZ"),
  rownames(tg)
)

if (length(hb_genes) > 0L) {
  tg[["percent.hb"]] <- PercentageFeatureSet(
    tg,
    features = hb_genes
  )
} else {
  tg$percent.hb <- 0
}

qc_metric_comparison <- FeatureScatter(
  tg,
  feature1 = "percent_mt_author",
  feature2 = "percent.mt"
) +
  ggtitle("Author-supplied vs recalculated mitochondrial percentage")

qc_metric_comparison
```

## QC distributions

```{r}
#| label: qc-distributions
#| fig-width: 14
#| fig-height: 5

qc_by_donor <- VlnPlot(
  tg,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo"),
  group.by = "donor",
  pt.size = 0,
  ncol = 4
)

qc_by_donor
```

```{r}
#| label: qc-feature-count-relationship
#| fig-width: 12
#| fig-height: 5

feature_count_scatter <- FeatureScatter(
  tg,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA",
  group.by = "cell_type_author"
)

count_mt_scatter <- FeatureScatter(
  tg,
  feature1 = "nCount_RNA",
  feature2 = "percent.mt",
  group.by = "cell_type_author"
)

feature_count_scatter + count_mt_scatter
```

## Conservative QC flags

These are review flags only. Do not subset nuclei until their effects are
assessed by donor and author-defined subtype.

```{r}
#| label: qc-flags

qc_min_features <- 200L
qc_min_counts <- 300L
qc_max_mt <- 10

tg$qc_low_features <- tg$nFeature_RNA < qc_min_features
tg$qc_low_counts <- tg$nCount_RNA < qc_min_counts
tg$qc_high_mt <- tg$percent.mt > qc_max_mt

tg$qc_flag <- tg$qc_low_features |
  tg$qc_low_counts |
  tg$qc_high_mt

tg$qc_status <- factor(
  ifelse(tg$qc_flag, "flagged", "retain"),
  levels = c("retain", "flagged")
)
```

## Initial neuron subset

This subset preserves all author-defined neuronal nuclei for the first
neuron-focused QC and visualization pass. The original full TG object remains
the source for later broad cell-class reference analyses.

```{r}
#| label: neuron-subset

tg_neurons_raw <- subset(
  tg,
  subset = cell_type_author == "neuron"
)

neuron_subtype_manifest <- tg_neurons_raw[[]] |>
  count(donor, subtype_author, name = "n_nuclei") |>
  arrange(subtype_author, donor)

knitr::kable(neuron_subtype_manifest)

write.csv(
  neuron_subtype_manifest,
  "results/qc/GSE197289_neuron_donor_by_subtype_nuclei.csv",
  row.names = FALSE
)

tg_neurons_raw
```

## Neuronal normalization and visualization

This section uses SCTransform for neuron-focused visualization, PCA, UMAP, and
exploratory clustering. The raw `RNA` counts in `tg_neurons_raw` remain the
exclusive input for pseudobulk construction.

```{r}
#| label: neuron-sct-umap
#| cache: true

tg_neurons_sct <- SCTransform(
  tg_neurons_raw,
  assay = "RNA",
  vst.flavor = "v2",
  variable.features.n = 3000,
  verbose = TRUE
)

tg_neurons_sct <- RunPCA(
  tg_neurons_sct,
  assay = "SCT",
  npcs = 50,
  verbose = FALSE
)

ElbowPlot(tg_neurons_sct, ndims = 50)
```

```{r}
#| label: neuron-umap-cluster
#| cache: true
#| fig-width: 14
#| fig-height: 10

dims_use <- 1:30

tg_neurons_sct <- FindNeighbors(
  tg_neurons_sct,
  reduction = "pca",
  dims = dims_use,
  verbose = FALSE
)

tg_neurons_sct <- FindClusters(
  tg_neurons_sct,
  resolution = 0.4,
  cluster.name = "cluster_sct_r04",
  verbose = FALSE
)

tg_neurons_sct <- RunUMAP(
  tg_neurons_sct,
  reduction = "pca",
  dims = dims_use,
  reduction.name = "umap_sct",
  seed.use = 197289,
  verbose = FALSE
)


p_donor <- DimPlot(
  tg_neurons_sct,
  reduction = "umap_sct",
  group.by = "donor"
) +
  ggtitle("TG donor")


p_subtype <- DimPlot(
  tg_neurons_sct,
  reduction = "umap_sct",
  group.by = "subtype_author",
  label = TRUE,
  repel = TRUE
) +
  ggtitle("Author-defined human TG neuronal subtypes")

p_cluster <- DimPlot(
  tg_neurons_sct,
  reduction = "umap_sct",
  group.by = "cluster_sct_r04",
  label = TRUE,
  repel = TRUE
) +
  ggtitle("Exploratory Seurat clusters")

p_donor

p_subtype + p_cluster
```

```{r}
#| label: subtype-cluster-concordance

subtype_cluster_concordance <- tg_neurons_sct[[]] |>
  count(subtype_author, cluster_sct_r04, name = "n_nuclei") |>
  group_by(subtype_author) |>
  mutate(
    subtype_total = sum(n_nuclei),
    fraction_within_subtype = n_nuclei / subtype_total
  ) |>
  ungroup() |>
  arrange(subtype_author, desc(fraction_within_subtype))

knitr::kable(subtype_cluster_concordance, digits = 3)

write.csv(
  subtype_cluster_concordance,
  "results/qc/GSE197289_neuron_subtype_cluster_concordance.csv",
  row.names = FALSE
)
```

## Neuronal marker validation

```{r}
#| label: neuron-marker-plots
#| cache: true
#| fig-width: 14
#| fig-height: 14

marker_panel <- c(
  "RBFOX3", "SNAP25", "SYT1", "ELAVL4", "TUBB3",
  "NTRK1", "NTRK2", "NTRK3",
  "TRPV1", "CALCA", "TAC1",
  "P2RX3", "TRPM8", "SST"
)

marker_panel_present <- intersect(marker_panel, rownames(tg_neurons_sct))
marker_panel_missing <- setdiff(marker_panel, marker_panel_present)

message(
  "Present markers: ", paste(marker_panel_present, collapse = ", "),
  "\nMissing markers: ", paste(marker_panel_missing, collapse = ", ")
)

FeaturePlot(
  tg_neurons_sct,
  features = marker_panel_present,
  reduction = "umap_sct",
  min.cutoff = "q05",
  max.cutoff = "q95",
  ncol = 4
)
```
```{r}
#| label: marker-violin-plots
#| fig-width: 14
#| fig-height: 8

violin_marker_panel <- c(
  "NTRK1", "NTRK2", "NTRK3",
  "TRPV1", "CALCA", "TAC1",
  "MRGPRD", "P2RX3", "TRPM8", "SST"
)

violin_marker_present <- intersect(
  violin_marker_panel,
  rownames(tg_neurons_sct)
)

violin_marker_missing <- setdiff(
  violin_marker_panel,
  violin_marker_present
)

message(
  "Markers omitted because absent from matrix: ",
  paste(violin_marker_missing, collapse = ", ")
)

VlnPlot(
  tg_neurons_sct,
  features = violin_marker_present,
  group.by = "subtype_author",
  pt.size = 0,
  ncol = 5
)
```


## Donor-by-subtype raw-count pseudobulk

Pseudobulk profiles are sums of raw UMI counts across nuclei grouped by biological
donor and author-defined neuronal subtype. Only groups with at least 20 nuclei
are retained for bulk-RNA comparison.

```{r}
#| label: neuronal-pseudobulk-manifest

minimum_nuclei_per_group <- 20L

pb_manifest <- tg_neurons_raw[[]] |>
  transmute(
    donor = as.character(donor),
    subtype_author = as.character(subtype_author)
  ) |>
  count(donor, subtype_author, name = "n_nuclei") |>
  mutate(
    include_pseudobulk = n_nuclei >= minimum_nuclei_per_group,
    exclusion_reason = if_else(
      include_pseudobulk,
      NA_character_,
      paste0("< ", minimum_nuclei_per_group, " nuclei")
    ),
    pseudobulk_id = paste(donor, subtype_author, sep = "__")
  ) |>
  arrange(subtype_author, donor)

knitr::kable(pb_manifest)

write.csv(
  pb_manifest,
  "results/pseudobulk/GSE197289_neuronal_pseudobulk_manifest.csv",
  row.names = FALSE
)

```

```{r}
#| label: construct-raw-pseudobulk
#| cache: true

# Raw counts are retrieved explicitly from the unmodified RNA assay.
raw_counts <- GetAssayData(
  tg_neurons_raw,
  assay = "RNA",
  layer = "counts"
)

# Establish a cell-level grouping table in the exact matrix-column order.
cell_metadata <- tg_neurons_raw[[]] |>
  as.data.frame() |>
  mutate(
    barcode = rownames(tg_neurons_raw[[]]),
    donor = as.character(donor),
    subtype_author = as.character(subtype_author),
    pseudobulk_id = paste(donor, subtype_author, sep = "__")
  )

stopifnot(identical(cell_metadata$barcode, colnames(raw_counts)))
stopifnot(nrow(cell_metadata) == ncol(raw_counts))

# Keep only donor x subtype groups meeting the predeclared cell-count criterion.
pb_sample_metadata <- pb_manifest |>
  filter(include_pseudobulk) |>
  mutate(
    donor = as.character(donor),
    subtype_author = as.character(subtype_author),
    pseudobulk_id = as.character(pseudobulk_id)
  ) |>
  arrange(subtype_author, donor)

group_levels <- pb_sample_metadata$pseudobulk_id

# Map every nucleus to its retained pseudobulk group.
cell_group <- cell_metadata$pseudobulk_id
keep_cell <- cell_group %in% group_levels

message(
  "Nuclei retained for pseudobulk: ", sum(keep_cell), " / ", length(keep_cell),
  "\nPseudobulk groups retained: ", length(group_levels)
)

# Avoid a barcode-based two-dimensional matrix subset:
# use logical selection, which remains in exact count-matrix order.
raw_counts_keep <- raw_counts[, keep_cell]

cell_group_keep <- factor(
  cell_group[keep_cell],
  levels = group_levels
)

# Sparse cells x pseudobulk membership matrix.
membership <- Matrix::sparse.model.matrix(~ 0 + cell_group_keep)

# Matrix columns must follow pseudobulk metadata order.
colnames(membership) <- group_levels

stopifnot(nrow(membership) == ncol(raw_counts_keep))
stopifnot(identical(colnames(raw_counts_keep), cell_metadata$barcode[keep_cell]))

# genes x cells multiplied by cells x groups -> genes x pseudobulk samples.
pb_counts <- raw_counts_keep %*% membership

stopifnot(ncol(pb_counts) == nrow(pb_sample_metadata))
stopifnot(identical(colnames(pb_counts), pb_sample_metadata$pseudobulk_id))
stopifnot(all(pb_counts@x == floor(pb_counts@x)))

pb_library_sizes <- tibble(
  pseudobulk_id = colnames(pb_counts),
  total_umis = Matrix::colSums(pb_counts),
  detected_genes = Matrix::colSums(pb_counts > 0)
) |>
  left_join(pb_sample_metadata, by = "pseudobulk_id")

pb_library_sizes <- pb_library_sizes |>
  mutate(
    confidence_tier = case_when(
      n_nuclei >= 50 ~ "primary",
      n_nuclei >= 20 ~ "exploratory",
      TRUE ~ "excluded"
    ),
    confidence_tier = factor(
      confidence_tier,
      levels = c("primary", "exploratory", "excluded")
    ),
    mean_umis_per_nucleus = total_umis / n_nuclei,
    genes_per_1000_umis = 1000 * detected_genes / total_umis
  )

knitr::kable(
  pb_library_sizes |>
    arrange(confidence_tier, subtype_author, donor),
  digits = 2
)
saveRDS(
  pb_counts,
  "results/pseudobulk/GSE197289_TG_neuron_donor_subtype_raw_counts.rds"
)

write.csv(
  pb_sample_metadata,
  "results/pseudobulk/GSE197289_TG_neuron_donor_subtype_metadata.csv",
  row.names = FALSE
)

write.csv(
  pb_library_sizes,
  "results/pseudobulk/GSE197289_TG_neuron_donor_subtype_library_sizes.csv",
  row.names = FALSE
)

tibble(
  metric = c(
    "Genes",
    "Eligible donor-by-subtype pseudobulks",
    "Nuclei represented",
    "Nuclei excluded by minimum-group rule"
  ),
  value = c(
    nrow(pb_counts),
    ncol(pb_counts),
    sum(keep_cell),
    sum(!keep_cell)
  )
) |>
  knitr::kable()
```

## Pseudobulk normalization and reference PCA

```{r}
#| label: pseudobulk-edger
#| cache: true

pb_dge <- DGEList(
  counts = pb_counts,
  samples = as.data.frame(pb_sample_metadata)
)

# For descriptive reference comparisons, retain genes expressed in at least
# two pseudobulk samples. Do not use subtype as the filter group because some
# subtype groups have only two donor-level profiles.
keep_genes <- rowSums(cpm(pb_dge) >= 1) >= 2

pb_dge <- pb_dge[keep_genes, , keep.lib.sizes = FALSE]
pb_dge <- calcNormFactors(pb_dge, method = "TMM")

pb_logcpm <- cpm(
  pb_dge,
  log = TRUE,
  prior.count = 2
)

stopifnot(identical(colnames(pb_logcpm), pb_sample_metadata$pseudobulk_id))

saveRDS(
  pb_dge,
  "results/pseudobulk/GSE197289_TG_neuron_donor_subtype_edgeR_dge.rds"
)

write.csv(
  pb_logcpm,
  "results/pseudobulk/GSE197289_TG_neuron_donor_subtype_logCPM.csv",
  row.names = TRUE
)

tibble(
  metric = c(
    "Pseudobulk samples",
    "Genes before expression filter",
    "Genes retained",
    "Median raw library size"
  ),
  value = c(
    ncol(pb_counts),
    nrow(pb_counts),
    nrow(pb_dge),
    median(pb_dge$samples$lib.size)
  )
) |>
  knitr::kable()
```

## Neuronal subtype marker heatmap

This heatmap displays TMM-normalized pseudobulk expression for curated neuronal
and sensory-subtype markers. Values are gene-wise z-scores of log-CPM, so each
row emphasizes relative expression across donor-by-subtype pseudobulks.

```{r}
#| label: pseudobulk-marker-heatmap
#| fig-width: 13
#| fig-height: 12

marker_sets <- list(
  `Pan-neuronal` = c(
    "RBFOX3", "STMN2", "TUBB3", "ELAVL4",
    "SNAP25", "SYT1", "UCHL1", "PRPH"
  ),
  `Peptidergic / nociceptor` = c(
    "NTRK1", "TRPV1", "CALCA", "CALCB", "TAC1",
    "SCN9A", "SCN10A", "SCN11A", "ASIC3", "PIRT",
    "PRDM12", "POU4F1", "TLX3"
  ),
  `Non-peptidergic` = c(
    "P2RX3", "RET", "GFRA3", "MRGPRX1",
    "TAFA3", "IL17B", "PTGDR"
  ),
  `Neurofilament / mechanosensory` = c(
    "NEFH", "NEFM", "NEFL", "NTRK2", "NTRK3",
    "PIEZO2", "KCNS1", "KCNH6", "KCNIP3"
  ),
  `Cold / pruriceptor-associated` = c(
    "TRPM8", "SST", "NMB", "HTR3A", "SLC17A6"
  )
)

marker_table <- tibble(
  gene = unlist(marker_sets, use.names = FALSE),
  marker_group = rep(
    names(marker_sets),
    lengths(marker_sets)
  )
) |>
  distinct(gene, .keep_all = TRUE)

marker_table <- marker_table |>
  mutate(
    gene = toupper(gene),
    present = gene %in% rownames(pb_logcpm)
  )

marker_missing <- marker_table |>
  filter(!present)

marker_present <- marker_table |>
  filter(present)

message(
  "Markers retained: ", nrow(marker_present),
  " / ", nrow(marker_table),
  "\nAbsent markers: ",
  paste(marker_missing$gene, collapse = ", ")
)

marker_matrix <- pb_logcpm[
  marker_present$gene,
  pb_sample_metadata$pseudobulk_id,
  drop = FALSE
]

# Gene-wise z-score: display relative enrichment across pseudobulk profiles.
marker_matrix_z <- t(scale(t(marker_matrix)))

# Remove rows that cannot be scaled because expression is invariant.
marker_matrix_z <- marker_matrix_z[
  apply(marker_matrix_z, 1, function(x) all(is.finite(x))),
  ,
  drop = FALSE
]

marker_annotation <- pb_sample_metadata |>
  select(pseudobulk_id, donor, subtype_author, n_nuclei) |>
  as.data.frame()

rownames(marker_annotation) <- marker_annotation$pseudobulk_id
marker_annotation$pseudobulk_id <- NULL

marker_row_annotation <- marker_present |>
  filter(gene %in% rownames(marker_matrix_z)) |>
  select(gene, marker_group) |>
  as.data.frame()

rownames(marker_row_annotation) <- marker_row_annotation$gene
marker_row_annotation$gene <- NULL

pheatmap(
  marker_matrix_z,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  annotation_col = marker_annotation,
  annotation_row = marker_row_annotation,
  gaps_row = cumsum(
    table(marker_row_annotation$marker_group)
  )[-length(unique(marker_row_annotation$marker_group))],
  show_colnames = TRUE,
  show_rownames = TRUE,
  fontsize_row = 9,
  fontsize_col = 8,
  border_color = NA,
  main = "Human TG neuronal pseudobulk marker expression"
)

write.csv(
  marker_table,
  "results/pseudobulk/GSE197289_TG_marker_heatmap_gene_manifest.csv",
  row.names = FALSE
)
```


## Broad TG cell-type pseudobulk reference

This reference aggregates nuclei by donor and broad author-defined TG cell type.
It is intended for transcriptional purity assessment of bulk RNA-seq samples;
it is not a literal cell-fraction deconvolution model.

```{r}
#| label: define-tg-broad-cell-types

tg$tg_broad_type <- dplyr::case_when(
  tg$subtype_author %in% c(
    "NF1", "NF2", "NF3", "PEP",
    "NP", "SST", "TRPM8", "cLTMR"
  ) ~ "Neuron",

  tg$subtype_author == "Satglia" ~ "Satellite glia",

  tg$subtype_author %in% c(
    "Schwann M", "Schwann N"
  ) ~ "Schwann cell",

  tg$subtype_author %in% c(
    "Fibroblast Mgpt", "Fibroblast Dcnt", "Fibroblast Apod"
  ) ~ "Fibroblast",

  tg$subtype_author == "Vascular" ~ "Vascular",

  tg$subtype_author == "Immune" ~ "Immune",

  TRUE ~ "Other"
)

tg$tg_broad_type <- factor(
  tg$tg_broad_type,
  levels = c(
    "Neuron",
    "Satellite glia",
    "Schwann cell",
    "Fibroblast",
    "Vascular",
    "Immune",
    "Other"
  )
)

tg_broad_manifest <- tg[[]] |>
  as.data.frame() |>
  mutate(donor = as.character(donor)) |>
  count(donor, tg_broad_type, name = "n_nuclei") |>
  mutate(
    pseudobulk_id = paste(donor, tg_broad_type, sep = "__"),
    include_pseudobulk = n_nuclei >= 20L
  ) |>
  arrange(tg_broad_type, donor)

knitr::kable(tg_broad_manifest)

write.csv(
  tg_broad_manifest,
  "results/pseudobulk/GSE197289_TG_broad_type_manifest.csv",
  row.names = FALSE
)
```

```{r}
#| label: construct-tg-broad-pseudobulk
#| cache: true

tg_raw_counts <- GetAssayData(
  tg,
  assay = "RNA",
  layer = "counts"
)

tg_cell_metadata <- tg[[]] |>
  as.data.frame() |>
  mutate(
    barcode = rownames(tg[[]]),
    donor = as.character(donor),
    tg_broad_type = as.character(tg_broad_type),
    pseudobulk_id = paste(donor, tg_broad_type, sep = "__")
  )

stopifnot(identical(tg_cell_metadata$barcode, colnames(tg_raw_counts)))

tg_pb_metadata <- tg_broad_manifest |>
  filter(include_pseudobulk) |>
  mutate(
    donor = as.character(donor),
    tg_broad_type = as.character(tg_broad_type)
  ) |>
  arrange(tg_broad_type, donor)

tg_pb_ids <- tg_pb_metadata$pseudobulk_id
tg_cell_group <- tg_cell_metadata$pseudobulk_id
tg_keep_cell <- tg_cell_group %in% tg_pb_ids

tg_membership <- Matrix::sparse.model.matrix(
  ~ 0 + factor(
    tg_cell_group[tg_keep_cell],
    levels = tg_pb_ids
  )
)

colnames(tg_membership) <- tg_pb_ids

tg_broad_pb_counts <- tg_raw_counts[, tg_keep_cell, drop = FALSE] %*%
  tg_membership

stopifnot(identical(
  colnames(tg_broad_pb_counts),
  tg_pb_metadata$pseudobulk_id
))
stopifnot(all(tg_broad_pb_counts@x == floor(tg_broad_pb_counts@x)))

tg_broad_pb_qc <- tibble(
  pseudobulk_id = colnames(tg_broad_pb_counts),
  total_umis = Matrix::colSums(tg_broad_pb_counts),
  detected_genes = Matrix::colSums(tg_broad_pb_counts > 0)
) |>
  left_join(tg_pb_metadata, by = "pseudobulk_id") |>
  mutate(
    mean_umis_per_nucleus = total_umis / n_nuclei
  )

knitr::kable(tg_broad_pb_qc, digits = 2)

saveRDS(
  tg_broad_pb_counts,
  "results/pseudobulk/GSE197289_TG_broad_type_donor_raw_counts.rds"
)

write.csv(
  tg_broad_pb_qc,
  "results/pseudobulk/GSE197289_TG_broad_type_donor_metadata.csv",
  row.names = FALSE
)
```

```{r}
#| label: normalize-tg-broad-pseudobulk
#| cache: true

tg_broad_dge <- edgeR::DGEList(
  counts = tg_broad_pb_counts,
  samples = as.data.frame(tg_pb_metadata)
)

tg_broad_keep_genes <- rowSums(
  edgeR::cpm(tg_broad_dge) >= 1
) >= 2L

tg_broad_dge <- tg_broad_dge[
  tg_broad_keep_genes,
  ,
  keep.lib.sizes = FALSE
]

tg_broad_dge <- edgeR::calcNormFactors(
  tg_broad_dge,
  method = "TMM"
)

tg_broad_logcpm <- edgeR::cpm(
  tg_broad_dge,
  log = TRUE,
  prior.count = 2
)

write.csv(
  tg_broad_logcpm,
  "results/pseudobulk/GSE197289_TG_broad_type_donor_logCPM.csv",
  row.names = TRUE
)

saveRDS(
  tg_broad_dge,
  "results/pseudobulk/GSE197289_TG_broad_type_donor_edgeR_dge.rds"
)
```

## Broad TG marker programs

```{r}
#| label: tg-broad-purity-marker-heatmap
#| fig-width: 14
#| fig-height: 11

purity_markers <- list(
  `Pan-neuronal` = c(
    "RBFOX3", "SNAP25", "SYT1", "STMN2",
    "ELAVL4", "TUBB3", "UCHL1", "PRPH"
  ),
  `Sensory neuron` = c(
    "NTRK1", "NTRK2", "NTRK3", "SCN9A",
    "SCN10A", "P2RX3", "TRPV1", "CALCA"
  ),
  `Satellite glia` = c(
    "S100B", "SLC1A3", "GLUL", "KCNJ10",
    "FABP7", "PLP1", "SLC1A2"
  ),
  `Schwann cell` = c(
    "SOX10", "MPZ", "MBP", "PMP22",
    "MAG", "NGFR", "NCAM1"
  ),
  `Fibroblast` = c(
    "COL1A1", "COL1A2", "COL3A1", "DCN",
    "LUM", "COL6A1", "COL6A2", "PDGFRA"
  ),
  `Vascular` = c(
    "VWF", "KDR", "EMCN", "PECAM1",
    "RAMP2", "PLVAP", "CLDN5"
  ),
  `Immune` = c(
    "PTPRC", "TYROBP", "LST1", "AIF1",
    "C1QA", "C1QB", "C1QC", "HLA-DRA"
  )
)

purity_marker_table <- tibble(
  gene = unlist(purity_markers, use.names = FALSE),
  marker_group = rep(names(purity_markers), lengths(purity_markers))
) |>
  mutate(gene = toupper(gene)) |>
  distinct(gene, .keep_all = TRUE) |>
  filter(gene %in% rownames(tg_broad_logcpm))

purity_matrix <- tg_broad_logcpm[
  purity_marker_table$gene,
  tg_pb_metadata$pseudobulk_id,
  drop = FALSE
]

purity_matrix_z <- t(scale(t(purity_matrix)))

purity_matrix_z <- purity_matrix_z[
  apply(purity_matrix_z, 1, function(x) all(is.finite(x))),
  ,
  drop = FALSE
]

purity_col_annotation <- tg_pb_metadata |>
  select(pseudobulk_id, donor, tg_broad_type, n_nuclei) |>
  as.data.frame()

rownames(purity_col_annotation) <- purity_col_annotation$pseudobulk_id
purity_col_annotation$pseudobulk_id <- NULL

purity_row_annotation <- purity_marker_table |>
  filter(gene %in% rownames(purity_matrix_z)) |>
  select(gene, marker_group) |>
  as.data.frame()

rownames(purity_row_annotation) <- purity_row_annotation$gene
purity_row_annotation$gene <- NULL

pheatmap::pheatmap(
  purity_matrix_z,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  annotation_col = purity_col_annotation,
  annotation_row = purity_row_annotation,
  show_colnames = TRUE,
  show_rownames = TRUE,
  fontsize_row = 9,
  fontsize_col = 8,
  border_color = NA,
  main = "Human TG broad cell-type pseudobulk marker programs"
)

write.csv(
  purity_marker_table,
  "results/pseudobulk/GSE197289_TG_broad_purity_marker_manifest.csv",
  row.names = FALSE
)
```
## Combined pseudobulk reference

This integrated reference combines donor-by-neuronal-subtype pseudobulks with
donor-by-broad non-neuronal TG pseudobulks. One shared TMM normalization produces
a common expression scale for the integrated PCA and marker heatmap.

```{r}
#| label: build-combined-pseudobulk-reference
#| cache: false

stopifnot(identical(
  rownames(pb_counts),
  rownames(tg_broad_pb_counts)
))

stopifnot(!anyDuplicated(colnames(pb_counts)))
stopifnot(!anyDuplicated(colnames(tg_broad_pb_counts)))

# Exclude donor-level neuronal pseudobulks because donor-by-neuronal-subtype
# profiles provide the higher-resolution neuronal reference.
tg_non_neuronal_ids <- tg_pb_metadata |>
  filter(tg_broad_type != "Neuron") |>
  pull(pseudobulk_id)

combined_counts <- cbind(
  pb_counts,
  tg_broad_pb_counts[, tg_non_neuronal_ids, drop = FALSE]
)

combined_metadata_neuron <- pb_library_sizes |>
  transmute(
    pseudobulk_id,
    donor = as.character(donor),
    reference_level = "Neuronal subtype",
    reference_label = as.character(subtype_author),
    n_nuclei,
    total_umis,
    detected_genes
  )

combined_metadata_non_neuron <- tg_broad_pb_qc |>
  filter(pseudobulk_id %in% tg_non_neuronal_ids) |>
  transmute(
    pseudobulk_id,
    donor = as.character(donor),
    reference_level = "Broad non-neuronal",
    reference_label = as.character(tg_broad_type),
    n_nuclei,
    total_umis,
    detected_genes
  )

combined_metadata <- bind_rows(
  combined_metadata_neuron,
  combined_metadata_non_neuron
) |>
  mutate(
    reference_level = factor(
      reference_level,
      levels = c("Neuronal subtype", "Broad non-neuronal")
    ),
    reference_label = factor(reference_label)
  )

combined_metadata <- combined_metadata[
  match(colnames(combined_counts), combined_metadata$pseudobulk_id),
  ,
  drop = FALSE
]

stopifnot(identical(
  combined_metadata$pseudobulk_id,
  colnames(combined_counts)
))

stopifnot(!anyDuplicated(combined_metadata$pseudobulk_id))
stopifnot(all(combined_counts@x == floor(combined_counts@x)))

combined_dge <- edgeR::DGEList(
  counts = combined_counts,
  samples = as.data.frame(combined_metadata)
)

combined_keep_genes <- rowSums(
  edgeR::cpm(combined_dge) >= 1
) >= 2L

combined_dge <- combined_dge[
  combined_keep_genes,
  ,
  keep.lib.sizes = FALSE
]

combined_dge <- edgeR::calcNormFactors(
  combined_dge,
  method = "TMM"
)

combined_logcpm <- edgeR::cpm(
  combined_dge,
  log = TRUE,
  prior.count = 2
)

combined_metadata <- combined_metadata |>
  mutate(
    mean_umis_per_nucleus = total_umis / n_nuclei
  )

stopifnot(identical(
  colnames(combined_logcpm),
  combined_metadata$pseudobulk_id
))

combined_reference_summary <- tibble(
  metric = c(
    "Combined pseudobulk profiles",
    "Neuronal subtype profiles",
    "Broad non-neuronal profiles",
    "Genes before expression filter",
    "Genes retained"
  ),
  value = c(
    ncol(combined_counts),
    sum(combined_metadata$reference_level == "Neuronal subtype"),
    sum(combined_metadata$reference_level == "Broad non-neuronal"),
    nrow(combined_counts),
    nrow(combined_dge)
  )
)

knitr::kable(combined_reference_summary)

saveRDS(
  combined_dge,
  "results/pseudobulk/GSE197289_TG_combined_reference_edgeR_dge.rds"
)

write.csv(
  combined_metadata,
  "results/pseudobulk/GSE197289_TG_combined_reference_metadata.csv",
  row.names = FALSE
)

write.csv(
  combined_logcpm,
  "results/pseudobulk/GSE197289_TG_combined_reference_logCPM.csv",
  row.names = TRUE
)
```



```{r}
#| label: combined-pseudobulk-pca
#| fig-width: 11
#| fig-height: 8

combined_pca <- prcomp(
  t(combined_logcpm),
  center = TRUE,
  scale. = FALSE
)

combined_pca_var <- 100 * summary(combined_pca)$importance[2, 1:2]

combined_pca_df <- combined_metadata |>
  mutate(
    PC1 = combined_pca$x[pseudobulk_id, 1],
    PC2 = combined_pca$x[pseudobulk_id, 2]
  )

stopifnot(
  !anyNA(combined_pca_df$PC1),
  !anyNA(combined_pca_df$PC2)
)

ggplot(
  combined_pca_df,
  aes(
    x = PC1,
    y = PC2,
    color = reference_label,
    shape = reference_level,
    size = n_nuclei
  )
) +
  geom_point(alpha = 0.9) +
  ggrepel::geom_text_repel(
    aes(label = pseudobulk_id),
    size = 3,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_size_continuous(range = c(3, 9)) +
  labs(
    title = "Integrated human TG pseudobulk reference",
    subtitle = "Neuronal subtypes and broad non-neuronal TG cell classes",
    x = sprintf("PC1 (%.1f%%)", combined_pca_var[1]),
    y = sprintf("PC2 (%.1f%%)", combined_pca_var[2]),
    color = "Reference identity",
    shape = "Reference level",
    size = "Nuclei aggregated"
  ) +
  theme_classic()
```

## Integrated marker heatmap

```{r}
#| label: combined-pseudobulk-marker-heatmap
#| fig-width: 15
#| fig-height: 15

combined_marker_sets <- list(
  `Pan-neuronal` = c(
    "RBFOX3", "SNAP25", "SYT1", "STMN2",
    "ELAVL4", "TUBB3", "UCHL1", "PRPH"
  ),
  `Peptidergic nociceptor` = c(
    "NTRK1", "TRPV1", "CALCA", "CALCB",
    "TAC1", "SCN9A", "SCN10A", "SCN11A",
    "ASIC3", "PIRT", "PRDM12", "POU4F1", "TLX3"
  ),
  `Non-peptidergic nociceptor` = c(
    "P2RX3", "RET", "GFRA3", "MRGPRX1",
    "TAFA3", "IL17B", "PTGDR"
  ),
  `Neurofilament / mechanosensory` = c(
    "NEFH", "NEFM", "NEFL", "NTRK2",
    "NTRK3", "PIEZO2", "KCNS1", "KCNH6", "KCNIP3"
  ),
  `Cold / pruriceptor-associated` = c(
    "TRPM8", "SST", "NMB", "HTR3A", "SLC17A6"
  ),
  `Satellite glia` = c(
    "S100B", "SLC1A3", "GLUL", "KCNJ10",
    "FABP7", "PLP1", "SLC1A2"
  ),
  `Schwann cell` = c(
    "SOX10", "MPZ", "MBP", "PMP22",
    "MAG", "NGFR", "NCAM1"
  ),
  `Fibroblast` = c(
    "COL1A1", "COL1A2", "COL3A1", "DCN",
    "LUM", "COL6A1", "COL6A2", "PDGFRA"
  ),
  `Vascular` = c(
    "VWF", "KDR", "EMCN", "PECAM1",
    "RAMP2", "PLVAP", "CLDN5"
  ),
  `Immune` = c(
    "PTPRC", "TYROBP", "LST1", "AIF1",
    "C1QA", "C1QB", "C1QC", "HLA-DRA"
  )
)

combined_marker_manifest <- tibble(
  gene = unlist(combined_marker_sets, use.names = FALSE),
  marker_group = rep(
    names(combined_marker_sets),
    lengths(combined_marker_sets)
  )
) |>
  mutate(gene = toupper(gene)) |>
  distinct(gene, .keep_all = TRUE) |>
  mutate(present = gene %in% rownames(combined_logcpm))

combined_marker_missing <- combined_marker_manifest |>
  filter(!present)

combined_marker_manifest <- combined_marker_manifest |>
  filter(present)

message(
  "Markers retained: ", nrow(combined_marker_manifest),
  "\nAbsent markers: ",
  paste(combined_marker_missing$gene, collapse = ", ")
)

combined_marker_matrix <- combined_logcpm[
  combined_marker_manifest$gene,
  combined_metadata$pseudobulk_id,
  drop = FALSE
]

combined_marker_matrix_z <- t(scale(t(combined_marker_matrix)))

combined_marker_matrix_z <- combined_marker_matrix_z[
  apply(combined_marker_matrix_z, 1, function(x) all(is.finite(x))),
  ,
  drop = FALSE
]

combined_col_annotation <- combined_metadata |>
  select(
    pseudobulk_id,
    donor,
    reference_level,
    reference_label,
    n_nuclei
  ) |>
  as.data.frame()

rownames(combined_col_annotation) <- combined_col_annotation$pseudobulk_id
combined_col_annotation$pseudobulk_id <- NULL

combined_row_annotation <- combined_marker_manifest |>
  filter(gene %in% rownames(combined_marker_matrix_z)) |>
  select(gene, marker_group) |>
  as.data.frame()

rownames(combined_row_annotation) <- combined_row_annotation$gene
combined_row_annotation$gene <- NULL

pheatmap::pheatmap(
  combined_marker_matrix_z,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  annotation_col = combined_col_annotation,
  annotation_row = combined_row_annotation,
  show_colnames = TRUE,
  show_rownames = TRUE,
  fontsize_row = 8,
  fontsize_col = 7,
  border_color = NA,
  main = "Integrated human TG pseudobulk marker programs"
)

write.csv(
  combined_marker_manifest,
  "results/pseudobulk/GSE197289_TG_combined_marker_manifest.csv",
  row.names = FALSE
)
```

## Save final objects

```{r}
#| label: save-final-objects

saveRDS(
  tg,
  "results/seurat/GSE197289_human_TG_full_raw.rds"
)

saveRDS(
  tg_neurons_raw,
  "results/seurat/GSE197289_human_TG_neurons_raw.rds"
)

saveRDS(
  tg_neurons_sct,
  "results/seurat/GSE197289_human_TG_neurons_sct.rds"
)

final_outputs <- tibble(
  artifact = c(
    "Full TG Seurat object",
    "Raw neuronal Seurat object",
    "SCT neuronal Seurat object",
    "Neuronal pseudobulk raw counts",
    "Integrated TG pseudobulk edgeR object",
    "Integrated TG pseudobulk log-CPM"
  ),
  path = c(
    "results/seurat/GSE197289_human_TG_full_raw.rds",
    "results/seurat/GSE197289_human_TG_neurons_raw.rds",
    "results/seurat/GSE197289_human_TG_neurons_sct.rds",
    "results/pseudobulk/GSE197289_TG_neuron_donor_subtype_raw_counts.rds",
    "results/pseudobulk/GSE197289_TG_combined_reference_edgeR_dge.rds",
    "results/pseudobulk/GSE197289_TG_combined_reference_logCPM.csv"
  )
)

knitr::kable(final_outputs)
```
